---
title: "Analysis of Role of Dams on Recorded Dry Season Malaria Incidences in Kasungu"
output: 
  html_document:
    toc: true
    toc_float: true
latex_engine: MiKTeX
author: Clinton Nkolokosa
date: " Last edited `r format(Sys.time(),'%d %B %Y')`"
---

This geospatial statistical model uses routinely collected malaria case data, population data and remotely sensed data, such as open and vegetated water bodies, to estimate population living around open water bodies, expected malaria cases, and `standardised morbidity ratio` (`SMR`) of malaria. And ultimately, quantify the association between proximity to larval habitat and malaria risk in health facility catchment areas in Kasungu. The `SMR` compares the risk of morbidity in a population of interest with that of a standard population. In this case, our interest is to find out whether the number of dry season malaria cases in each catchment area are greater than we would expect given the malaria rate for the entire Kasungu district.

We do this by comparing what we *observe* (O) with what we would *expect* (E) if the risk of malaria was equal throughout Kasungu. The SMR statistical notation of catchment *i* can be written as follows: $$SMR_i = \frac{O_i}{E_i}$$

Buffers around waterbodies are created and then combined with population data in raster format to estimate the proprtion of catcment population living within 1km, 2km and 3km of water bodies. Subsequently, the observed malaria cases are modeled using `Poisson regression` to find out if living within various distances from water bodies is causing variability in malaria risk in Kasungu district. We hypothesize that the risk of being a case in a catchment is dependent on proximity to water bodies. The data used spans from 2017 to 2020 and was derived from digitized DHIS2 malaria records, accessibility mapping, aggregated population geospatial layer and TropWet tool in Google Earth Engine.

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', echo = TRUE)
```

## Load packages

Loading the R packages that will be used to read in, view, transform and model the malaria cases and spatial datasets.

```{r, warning = FALSE, message = FALSE}
library(SpatialEpi)
library(spdep)
library(spaMM)
library(popEpi)
library(Epi)
library(tidyverse)
library(caret)
library(stats)
library(performance)   # check linear model performance
library(see)           # to plot model assumptions
library(qqplotr)
library(ggpubr)
library(plotly)
library(lubridate)
library(knitr)
library(raster)
library(rgdal)
library(rgeos)
library(sf)
library(sp)
library(tmap)
library(spdep)
library(maptools)
library(gridExtra)
library(ggsci)
library(grid)
library(exactextractr)
library(DataExplorer)
library(mapview)
library(kableExtra)
`%>%` <- magrittr::`%>%`
```

## Tell R where the data is

```{r}
file.path(getwd(),"data")

here::here()
```

## Load datasets

The total dry season malaria cases recorded at health-care facilities in Kasungu from 2017 to 2019 are contained in the `KasunguData.csv` sourced from <https://dhis2.health.gov.mw/>. The `kasungu_facility_catchments_2004.shp` shapefile also contains the population and health information within each health-facility catchment area in Kasungu district.

The aggregated population raster layers for Malawi e.g.,`ku_pop_2017_1km_aggregated.tif` were downloaded from the Open Spatial and Demographic and Data Research website: <https://www.worldpop.org/geodata/country?iso3=MWI>. These layers estimate total number of people per grid-cell. The units are number of people per pixel with country totals adjusted to match the corresponding official United Nations population estimates. The datasets were downloaded in Geotiff at a resolution of 1km and are projected in Geographic Coordinate System, WGS84.

The `kasungu_water.shp`and `water_bodies` layers contain open and vegetated waterbodies polygons, detected using the Tropical Wetland Unmixing Tool (TropWet). TropWet is a Google Earth Engine hosted toolbox that uses the Landsat archive to map tropical wetlands and can be accessed through: <https://www.aber.ac.uk/en/dges/research/earth-observation-laboratory/research/tropwet/>

```{r, Kasungu water bodies polygons, malaria cases by health facility catchments and population layers, message= FALSE}

# Kasungu dry season malaria data
dry_season_malaria_2017_2020 <- read.csv(here::here("data/dry_season_malaria_2017_2020.csv"))

# Kasungu district boundary shapefile 
kasungu_district <- sf::st_read(here::here("data", "kasungu_district.shp"))
  
# Kasungu health facility catchments generated from accessibility mapping
malire_new <- sf::st_read(here::here("data", "new_catchments.shp")) |> 
              sf::st_transform(32736) # reproject to WGS UTM Zone 36 South

# Kasungu population raster layer
kasungu_population_2017 <- raster(here::here("data", "ku_pop_2017_1km_aggregated.tif"))

kasungu_population_2018 <- raster(here::here("data", "ku_pop_2018_1km_aggregated.tif"))

kasungu_population_2019 <- raster(here::here("data", "ku_pop_2019_1km_aggregated.tif"))

kasungu_population_2020 <- raster(here::here("data", "ku_pop_2020_1km_aggregated.tif"))

# Read in waterbodies polygons 
dryseason_waterbodies_2017 <- sf::st_read(here::here("data", "water_bodies_2017.shp"))

dryseason_waterbodies_2018 <- sf::st_read(here::here("data", "kasungu_2018_water.shp"))

dryseason_waterbodies_2019 <- sf::st_read(here::here("data", "kasungu_2019_water.shp"))

dryseason_waterbodies_2020 <- sf::st_read(here::here("data", "water_bodies_2020.shp"))

# Add a row ID to water bodies polygons 
dryseason_waterbodies_2017$ID <- 1:nrow(dryseason_waterbodies_2017)

dryseason_waterbodies_2018$ID <- 1:nrow(dryseason_waterbodies_2018)

dryseason_waterbodies_2019$ID <- 1:nrow(dryseason_waterbodies_2019)

dryseason_waterbodies_2020$ID <- 1:nrow(dryseason_waterbodies_2020)

```

## View the dry season malaria case data
Lets have a closer look at the malaria dataset.
We observe that Kasungu district has 30 health facilities classified as dispensary, health centre, district hospital and rural hospital, and the highest malaria cases were recorded at Kasungu District Hospital.

```{r, warning = FALSE, message = FALSE, fig.height = 10, fig.width = 8, fig.cap = 'Fig.1 The total malaria cases recorded at each health-care facility in Kasungu district'}

dry_season_malaria_2017_2020 |>
  summary()

# Plotly bar chart -------------------------------------------------------------
bar_chart <- dry_season_malaria_2017_2020 |>  
  dplyr::filter(Names != "K2 Taso Clinic",          # Have missing malaria records
                Names != "Kalikeni Private Clinic",
                Names != "Kakwale Health Centre",
                Names != "St Andrews Community Hospital",
                Names != "St. Faith Health Centre",
                Names != "Chambwe Health Centre") |> 
  plotly::plot_ly(y = ~Names,
                  x = ~dr_2017,
                  type = "bar",
                  orientation = 'h',
                  name = "2017") |>
  plotly::add_trace(x = ~ dr_2018,
                    name = "2018") |>
  plotly::add_trace(x = ~ dr_2019,
                    name = "2019") |>
  plotly::add_trace(x = ~ dr_2020,
                    name = "2020") |> 
  plotly::layout(xaxis = list(title = "Total malaria cases"),
                 yaxis = list(title = " "),
                 hovermode = "compare",
                 margin = list(b = 10,
                               t = 10,
                               pad = 2))
bar_chart

# Pivot longer -----------------------------------------------------------------
# dry_season_malaria_longer <- dry_season_malaria_2017_2020 |>
#   dplyr::filter(Names != "K2 Taso Clinic",                       
#                 Names != "Kalikeni Private Clinic",
#                 Names != "Kakwale Health Centre",
#                 Names != "St Andrews Community Hospital",
#                 Names != "St. Faith Health Centre",
#                 Names != "Chambwe Health Centre") |>
#   dplyr::rename(`2017` = dr_2017,
#                 `2018` = dr_2018,
#                 `2019` = dr_2019,
#                 `2020` = dr_2020) |>
#   tidyr::pivot_longer(cols = `2017`:`2020`,
#                       names_to = 'year',
#                       values_to = 'malaria_cases')
# 
# ggplot2::ggplot(dry_season_malaria_longer, 
#                 aes(x = malaria_cases, 
#                     y = Names,
#                     fill = year))+
#   ggplot2::geom_bar(stat ='identity',
#                     position = "dodge")+
#   ggplot2::labs(x = "Dry season malaria cases",
#                 y = " ")+
#   ggplot2::theme_classic()+
#   ggsci::scale_fill_jama()


```

## Kasungu health-care facilities and their catchment areas

Heath facility catchment area is the area from which a health facility attracts patients. The new health facility catchments polygon was generated from generic accessibility mapping script adapted from <https://malariaatlas.org/wp-content/uploads/accessibility/R_generic_accessibilty_mapping_script.r> The script requires two user supplied datasets: the `2015 friction surface`, which is available here: <http://www.map.ox.ac.uk/accessibility_to_cities/>, and a user-supplied .csv of points `dry_season_malaria_2017_2020`. The accumulated cost algorithm `accCost` and `r.Cost` algorithm in QGIS were run to make the final output map of new health facility catchment boundaries.

```{r, warning = FALSE, message = FALSE}

# Using the complete.cases() function to select health centres with complete 
# longitude and latitude coordinates.
zipatala_aggregated <- dry_season_malaria_2017_2020[complete.cases(dry_season_malaria_2017_2020),] 

# Aggregate health facilities close to each other: 
#  a) Kasalika Health Centre and Kasungu District Hospital, 
#  b) Bua and Mziza Health Centres, and 
#  c) Kaluluma and Nkhamenya Rural Hospitals in order to 
# generate catchment areas that are geographically correct

zipatala_aggregated$dr_2017[which(
  zipatala_aggregated$Names == "Kasungu District Hospital")] <- zipatala_aggregated$dr_2017[which(
    zipatala_aggregated$Names == "Kasungu District Hospital")] + zipatala_aggregated$dr_2017[which(
      zipatala_aggregated$Names == "Kasalika Health Centre")]

zipatala_aggregated$dr_2018[which(
  zipatala_aggregated$Names == "Kasungu District Hospital")] <- zipatala_aggregated$dr_2018[which(
    zipatala_aggregated$Names == "Kasungu District Hospital")] + zipatala_aggregated$dr_2018[which(
      zipatala_aggregated$Names == "Kasalika Health Centre")]

zipatala_aggregated$dr_2019[which(
  zipatala_aggregated$Names == "Kasungu District Hospital")] <- zipatala_aggregated$dr_2019[which(
    zipatala_aggregated$Names == "Kasungu District Hospital")] + zipatala_aggregated$dr_2019[which(
      zipatala_aggregated$Names == "Kasalika Health Centre")]

zipatala_aggregated$dr_2020[which(
  zipatala_aggregated$Names == "Kasungu District Hospital")] <- zipatala_aggregated$dr_2020[which(
    zipatala_aggregated$Names == "Kasungu District Hospital")] + zipatala_aggregated$dr_2020[which(
      zipatala_aggregated$Names == "Kasalika Health Centre")]
  
zipatala_aggregated$dr_2017[which(
  zipatala_aggregated$Names == "Nkhamenya Rural Hospital")] <- zipatala_aggregated$dr_2017[which(
    zipatala_aggregated$Names == "Nkhamenya Rural Hospital")] + zipatala_aggregated$dr_2017[which(
      zipatala_aggregated$Names == "Kaluluma Rural Hospital")]

zipatala_aggregated$dr_2018[which(
  zipatala_aggregated$Names == "Nkhamenya Rural Hospital")] <- zipatala_aggregated$dr_2018[which(
    zipatala_aggregated$Names == "Nkhamenya Rural Hospital")] +zipatala_aggregated$dr_2018[which(
      zipatala_aggregated$Names == "Kaluluma Rural Hospital")]

zipatala_aggregated$dr_2019[which(
  zipatala_aggregated$Names == "Nkhamenya Rural Hospital")] <- zipatala_aggregated$dr_2019[which(
    zipatala_aggregated$Names == "Nkhamenya Rural Hospital")] + zipatala_aggregated$dr_2019[which(
      zipatala_aggregated$Names == "Kaluluma Rural Hospital")]

zipatala_aggregated$dr_2020[which(
  zipatala_aggregated$Names == "Nkhamenya Rural Hospital")] <- zipatala_aggregated$dr_2020[which(
    zipatala_aggregated$Names == "Nkhamenya Rural Hospital")] + zipatala_aggregated$dr_2020[which(
      zipatala_aggregated$Names == "Kaluluma Rural Hospital")]

zipatala_aggregated$dr_2017[which(
  zipatala_aggregated$Names == "Mziza Health Centre")] <- zipatala_aggregated$dr_2017[which(
    zipatala_aggregated$Names == "Mziza Health Centre")] + zipatala_aggregated$dr_2017[which(
      zipatala_aggregated$Names == "Bua Health Centre")]

zipatala_aggregated$dr_2018[which(
  zipatala_aggregated$Names == "Mziza Health Centre")] <- zipatala_aggregated$dr_2018[which(
    zipatala_aggregated$Names == "Mziza Health Centre")] + zipatala_aggregated$dr_2018[which(
      zipatala_aggregated$Names == "Bua Health Centre")]

zipatala_aggregated$dr_2019[which(
  zipatala_aggregated$Names == "Mziza Health Centre")] <- zipatala_aggregated$dr_2019[which(
    zipatala_aggregated$Names == "Mziza Health Centre")] + zipatala_aggregated$dr_2019[which(
      zipatala_aggregated$Names == "Bua Health Centre")] 
  

zipatala_aggregated$dr_2020[which(
  zipatala_aggregated$Names == "Mziza Health Centre")] <- zipatala_aggregated$dr_2020[which(
    zipatala_aggregated$Names == "Mziza Health Centre")] + zipatala_aggregated$dr_2020[which(
      zipatala_aggregated$Names == "Bua Health Centre")]

# Drop out the other health facilities
zipatala_aggregated <- zipatala_aggregated |>
  dplyr::filter(Names != "Kasalika Health Centre",
                Names != "Bua Health Centre",
                Names != "Kaluluma Rural Hospital")

# write.csv(zipatala_aggregated, "data/health_facilities_aggregated.csv")

# Convert to csv spatial to spatial object
health_facility_aggr_sf <- sf::st_as_sf(zipatala_aggregated,
                                        coords = c("LONGITU", "LATITUD"),
                                        crs = 4326, agr = "constant")

# st_write(health_facility_aggr_sf, "data/health_facilities_aggregated.shp")

```

View location of the health facilities in the new catchment areas

```{r, message=FALSE, fig.cap='Fig 2. Kasungu health-care facilities and catchment areas'}

# Plot map
tm_shape(malire_new)+
  tm_polygons()+
  tm_shape(health_facility_aggr_sf)+
  tm_dots(size = .3, 
          col = "blue", 
          alpha = 0.5)+
  tm_text("Names", 
          size = .3, 
          just = "top", 
          col = "black", 
          remove.overlap = TRUE)+
  tm_layout(frame = FALSE,
            title = "New Kasungu health facility \n catchment boundaries",
            title.size = .8, 
            title.position = c("left", "top"))+
  tm_compass(position=c("right", "top"))+
  tm_scale_bar(breaks = c(0, 10, 20), 
               text.size = .5)

```

## Kasungu district estimated population per grid-cell

```{r, message = FALSE, warning = FALSE, fig.height = 9, fig.width = 10, fig.cap = 'Fig.3 Estimated total number of people per 1km grid-cell'}

# Take a glimpse at the WorldPop raster layers
kasungu_population_2017

kasungu_population_2018

kasungu_population_2019

kasungu_population_2020

# Helper function to create a raster population map
create.population.map <- function(population.raster, title){
  # raster population map
  # arguments:
  #   population.raster:  aggregated population raster layer from WorldPop
  #   legend.title: legend title
  # returns:
  #   a tmap-element (plots a map)
  
  # set to interactive view
  # tmap::tmap_mode("view")
  
  # plot map
  tmap::tm_shape(population.raster)+
    tmap::tm_raster(palette = "-viridis", 
                    title = title,
                    breaks = c(0,100,200,400,600,800,1000,2000,4000,6000,8000))+
    tmap::tm_layout(legend.position = c("right", "bottom"),
                    frame = FALSE)+
    tmap::tm_scale_bar(position = c("left", "bottom"))
}
# Set to static map
 tmap_mode("plot")

estimated_pop_2017 <- create.population.map(kasungu_population_2017, title = "2017 Population")

estimated_pop_2018 <- create.population.map(kasungu_population_2018, title = "2018 Population")

estimated_pop_2019 <- create.population.map(kasungu_population_2019, title = "2019 Population")

estimated_pop_2020 <- create.population.map(kasungu_population_2020, title = "2020 Population")

# Layout the maps
tmap::tmap_arrange(estimated_pop_2017, estimated_pop_2018, 
                   estimated_pop_2019, estimated_pop_2020, nrow = 2) 

```

## Assign dry season malaria cases and population density to new health facility catchments

The WorldPop `aggregated population` e.g. `kasungu_population_2017.tif`, and DHIS2 malaria `dry_season_malaria_2017_2020` datasets are assigned to the new health facility catchments.

```{r, message=FALSE, warning=FALSE}
# Helper function that assigns malaria data from health facilities to their catchments areas ----------------
assign.malaria.data <- function(catchment_boundary, malaria_data){
  # arguments:
  #   catchment_boundary: sf polygon object of new catchment boundaries
  #   malaria_data: sf point object with a data frame containing the dry season malaria cases
  # returns:
  #   catchments_malaria_sf: sf polygon object with a data frame containing dry season malaria cases


  # Convert sf objects to spatial
  catchment_shp <- as(catchment_boundary, "Spatial")
  
  malaria_shp <- as(malaria_data, "Spatial")

  # Match CRS
  malaria_shp <- spTransform(malaria_shp, crs(catchment_shp))

  # Overlay aggregated health facility points and extract 2017 - 2020 malaria cases
  # Using 'point.in.poly' to return a point spatial object, in this case location of health facilities
  # and estimated population instead of sp::over function, which simply returns 
  # a data frame, with the same no. rows.
  # Argument 'sp = TRUE' returns an sp class object, else returns sf class object
  # Joining the malaria and population dataset using only 'merge' function can't work due to 
  # non-unique columns and differences in row numbers
  
  hospitals_in_catchment <- spatialEco::point.in.poly(malaria_shp, catchment_shp, sp = TRUE) 

  # Add the extracted ID, health facility names and dry season malaria cases to 
  # the health facility catchments (hfc)
  hfc_malaria_shp <- merge(catchment_shp, hospitals_in_catchment, by.x = "DN", by.y = "rowID")

  # Convert the shapefile containing malaria data to sf-object
  hfc_malaria_sf <- sf::st_as_sf(hfc_malaria_shp)

  # Tidy the data by dropping columns not needed
  catchment_malaria <- hfc_malaria_sf |> 
    dplyr::select(-c(coords.x1, coords.x2))

  return(out = catchment_malaria)
}


# Invoking the function ----------------------------------------------------------------------------------
malaria_by_catchment <- assign.malaria.data(malire_new, health_facility_aggr_sf)

```

Assign population data to the health catchment areas

```{r, message=FALSE}

# Helper unction to extract population from WorldPop raster file and assign ---------------------------
# the values to the new catchments.

extract.pop.values <- function(kasungu_pop_raster, catchments){
  # function to extract population from raster file and assign the population to catchments
  # arguments:
  #   kasungu_pop_raster: population raster file clipped to Kasungu district
  #   catchments: shapefile containing the polygons that we wish to use as boundaries
  # returns:
  #   catchments_malaria_pop_sf: sf polygon object containing malaria and population data
  
  # convert from sf to sp
  catchments_sp <- as(catchments, "Spatial")
  
  # Match extent i.e projection
  catchments_sp <- spTransform(catchments_sp, proj4string(kasungu_pop_raster))
  
  # Crop and mask the population raster to exclude Kasungu National Park
  pop_raster_clip <- raster::crop(kasungu_pop_raster, extent(catchments_sp)) |>
    raster::mask(catchments_sp)
  # pop_raster_clip <- raster::mask(raster::crop(kasungu_pop_raster, extent(catchments_sp)), catchments_sp)
  
  # Extracting zonal statistics from a population raster layer. 
  # The population raster is a continuous gridded surface layer that has an 
  # estimated population density value to every square in their grid. 
  # The population values are then summed and apportioned to the catchment polygons
  # catchments_malaria_pop <- catchments %>% 
  #   dplyr::mutate(pop = round(raster::extract(pop_raster_clip, catchments, fun = sum, na.rm = TRUE)))
  
  pop_by_catchment <- round(raster::extract(pop_raster_clip, catchments, fun = sum, na.rm = TRUE))
  
  pop_by_catchment_df <-  pop_by_catchment %>%  
  # apply unlist to the lists to have vectors as the list elements
  lapply(unlist) %>% 
  # convert vectors to data.frames
  lapply(as_tibble) %>% 
  # combine the list of data.frames
  bind_rows(., .id = "rowID") %>% 
  # rename the value variable
  dplyr::rename(pop = value)
  
  # Add row ID to column to catchment layer
  catchments$rowID <- 1:nrow(catchments)
  
  # Merge catchment areas and population data 
  pop_by_catchments <- merge(catchments, pop_by_catchment_df, by = "rowID")
  
  # Cleaning 'Inf' values
  pop_by_catchments %>% 
    dplyr::mutate_if(is.numeric, list(~na_if(., Inf))) %>% 
    dplyr::mutate_if(is.numeric, list(~na_if(., -Inf)))

  return(out = pop_by_catchments)
  
}

# Invoking the function ---------------------------------------------------------------------------------------
malaria_pop_by_catchment_2017 <- extract.pop.values(kasungu_population_2017, malaria_by_catchment)

malaria_pop_by_catchment_2018 <- extract.pop.values(kasungu_population_2018, malaria_by_catchment)

malaria_pop_by_catchment_2019 <- extract.pop.values(kasungu_population_2019, malaria_by_catchment)

malaria_pop_by_catchment_2020 <- extract.pop.values(kasungu_population_2020, malaria_by_catchment)


```

## View population by catchment maps

Estimated total number of people within health facility catchment areas.

```{r, message=FALSE, fig.height = 5, fig.width = 9, fig.cap = 'Fig. 4: Estimated population by health facility catchment areas'}
# Helper function to create maps of estimated population by catchment areas --------------------------------
create.population.map <- function(catchment.area, 
                                  variable = "pop", 
                                  title, 
                                  legend.title = "Estimated \n population"){
  # estimated population map
  # catchment.area: estimated population layer from nachulu function
  # variable: variable name (as character, in qoutes)
  # title: map title in quotes
  # legend.title: legend title in qoutes
  # returns:
  #   a tmap-element (plots a map)
  tm_shape(catchment.area)+
    tm_fill(col = variable, 
            breaks = c(0, 13000, 19000, 27000, 35000, 70000, 140000, 200000),
            palette = "YlOrBr",
            title = legend.title)+
    tm_borders(col = "grey",
               lwd = 0.4)+
    tm_layout(legend.position = c(0.75, "bottom"),
              legend.text.size = 0.6,
              legend.title.size = 0.8,
              frame = FALSE)+
    tm_credits(title, 
               position = c(0.3, 0.8), 
               size = 1)
}

# Invoking the function --------------------------------------------------------------------------------
pop_by_catchment_2017 <- create.population.map(malaria_pop_by_catchment_2017, title = "2017")

pop_by_catchment_2018 <- create.population.map(malaria_pop_by_catchment_2018, title = "2018")

pop_by_catchment_2019 <- create.population.map(malaria_pop_by_catchment_2019, title = "2019")

pop_by_catchment_2020 <- create.population.map(malaria_pop_by_catchment_2020, title = "2020")

tmap::tmap_arrange(pop_by_catchment_2017, pop_by_catchment_2018,
                   pop_by_catchment_2019, pop_by_catchment_2020, ncol = 2)


```

Population density by catchment

```{r, message=FALSE, fig.height = 5, fig.width = 9, fig.cap = 'Fig. 5: Estimated population density by health facility catchment areas'}
# Helper function to calculate population density by catchment -----------------
calculate.population.density <- function(pop.data){
  
  # Convert to spatial object
  pop.sp <- as(pop.data, "Spatial")

  # Calculate area of catchment polygon in square kilometres
  pop.sp$area_sqkm <- round(rgeos::gArea(pop.sp, byid = TRUE) / (1000 * 1000))

  # Calculate population density
  pop.sp$pop_density <- round(pop.sp$pop / pop.sp$area_sqkm)
  
  # Convert back to sf object
  pop.sf <- sf::st_as_sf(pop.sp)
  
  return(pop.sf)
}

# Invoking function ------------------------------------------------------------
pop_density_2017 <- calculate.population.density(malaria_pop_by_catchment_2017)

pop_density_2018 <- calculate.population.density(malaria_pop_by_catchment_2018)

pop_density_2019 <- calculate.population.density(malaria_pop_by_catchment_2019)

pop_density_2020 <- calculate.population.density(malaria_pop_by_catchment_2020)

# Helper function to create population density maps ----------------------------
create.pop.density.map <- function(pop.density.data,
                                   variable = "pop_density", 
                                   title = NA, 
                                   legend.title = "Population \ndensity/km^2"){
  tm_shape(pop.density.data)+
    tm_fill(col = variable, 
            breaks = c(0, 50, 100, 150, 200, 250, 300, 350),
            palette = "-magma",
            title = legend.title)+
    tm_borders()+
    tm_layout(legend.position = c(0.75, "bottom"),
              legend.text.size = 0.6,
              legend.title.size = 0.8,
              frame = FALSE)+
    tm_credits(title, 
               position = c(0.3, 0.8), 
               size = 1)
}

# Invoking function ------------------------------------------------------------
pop_density_2017_map <- create.pop.density.map(pop_density_2017, title = "2017")

pop_density_2018_map <- create.pop.density.map(pop_density_2018, title = "2018")

pop_density_2019_map <- create.pop.density.map(pop_density_2019, title = "2019")

pop_density_2020_map <- create.pop.density.map(pop_density_2020, title = "2020")

# Layout maps
tmap::tmap_arrange(pop_density_2017_map, pop_density_2018_map,
                   pop_density_2019_map, pop_density_2020_map, ncol = 2)

```

## Calculate the expected number of cases for each catchment area

The `expected` number of dry season malaria cases in catchment *i* are calculated as the observed risk (r) of malaria i.e. the total number of malaria cases in Kasungu district divided by the total population of the district, multiplied by the number of people in the catchment area: $$E_i = \frac{\sum_i O_i}{\sum_i N_i}\times N_i$$

The `expected` number of dry season malaria cases are calculated under the assumption that there is no spatial variation in risk, i.e., no difference in infection rates between the catchment areas.

```{r, message=FALSE}

# Calculate expected malaria cases --------------------------------------------------------------
expected_malaria_2017 <- malaria_pop_by_catchment_2017 |>
  dplyr::rename(
    observed_2017 = dr_2017,
     pop_2017 = pop) |> 
  dplyr::mutate(
    expected_2017 = round(sum(observed_2017)/sum(pop_2017, na.rm = TRUE)*pop_2017))

expected_malaria_2018 <- malaria_pop_by_catchment_2018 |>
  dplyr::rename(
    observed_2018 = dr_2018,
    pop_2018 = pop) |> 
  dplyr::mutate(
    expected_2018 = round(sum(observed_2018)/sum(pop_2018, na.rm = TRUE)*pop_2018))

expected_malaria_2019 <- malaria_pop_by_catchment_2019 |>
  dplyr::rename(
    observed_2019 = dr_2019,
    pop_2019 = pop) |>
  dplyr::mutate(
    expected_2019 = round(sum(observed_2019)/sum(pop_2019, na.rm = TRUE)*pop_2019)) 

expected_malaria_2020 <- malaria_pop_by_catchment_2020 |>
  dplyr::rename(
    observed_2020 = dr_2020,
    pop_2020 = pop) |>
  dplyr::mutate(
    expected_2020 = round(sum(observed_2020)/sum(pop_2020, na.rm = TRUE)*pop_2020))

```

## Calculate the Standardised Morbidity Ratio of malaria incidences for each catchment area

The `SMR` compares the risk of morbidity in a population of interest with that of a standard population. In this case, our interest is to find out whether the number of dry season malaria cases in each catchment area are greater than we would expect given the malaria rate for the entire Kasungu district.

We do this by comparing what we `observe` (O) with what we would `expect` (E) if the risk of malaria was equal throughout Kasungu. The SMR of catchment *i* can be calculated as follows: $$SMR_i = \frac{O_i}{E_i}$$

```{r}
# Calculate Standardised Morbidity Ratio (SMR) -------------------------------------
SMR_2017 <- expected_malaria_2017 |>
  dplyr::mutate(SMR = round(observed_2017/expected_2017, 1)) |> 
  dplyr::select(rowID,Names, pop_2017, observed_2017, expected_2017, SMR) 

SMR_2018 <- expected_malaria_2018 |> 
  dplyr::mutate(SMR = round(observed_2018/expected_2018, 1)) |> 
  dplyr::select(rowID, Names, pop_2018, observed_2018, expected_2018, SMR) 

SMR_2019 <- expected_malaria_2019 |> 
  dplyr::mutate(SMR = round(observed_2019/expected_2019, 1)) |> 
  dplyr::select(rowID, Names, pop_2019, observed_2019, expected_2019, SMR) 

SMR_2020 <- expected_malaria_2020 |> 
  dplyr::mutate(SMR = round(observed_2020/expected_2020, 1)) |> 
  dplyr::select(rowID, Names, pop_2020, observed_2020, expected_2020, SMR)


# Create SMR tables ------------------------------------------------------------
SMR_table_2017 <- SMR_2017 |>
  dplyr::as_tibble() |>
  dplyr::select(-rowID, -geometry) |> 
  kable() |>
  kableExtra::kable_styling(full_width = FALSE)

SMR_table_2018 <- SMR_2018 |> 
  dplyr::as_tibble() |> 
  dplyr::select(-rowID, -geometry) |>
  kable() |> 
  kableExtra::kable_styling(full_width = FALSE)

SMR_table_2019 <- SMR_2019 |>
  dplyr::as_tibble() |>
  dplyr::select(-rowID, -geometry) |> 
  kable () |> 
  kableExtra::kable_styling(full_width = FALSE)

SMR_table_2020 <- SMR_2020 |>
  dplyr::as_tibble() |> 
  dplyr::select(-rowID, -geometry) |> 
  kable () |> 
  kableExtra::kable_styling(full_width = FALSE)

SMR_table_2017

SMR_table_2018

SMR_table_2019

SMR_table_2020

```

## View observed and expected dry season malaria cases

```{r, message=FALSE, fig.height=10, fig.width=7, fig.cap= 'Fig 6: Observed and expected malaria incidence by health facility catchment area, Kasungu'}
# Helper function to create maps of observed and expected dry season malaria cases
create.malaria.map <- function(malaria.data, 
                               variable = NA, 
                               title = NA, 
                               legend.title = NA){
  # observed and expected malaria incidence map
  # malaria.data: data frame containing observed and expected malaria cases
  # variable: variable name (as character, in quotes e.g. "observed")
  # title: map title in quotes
  # legend.title: legend title in quotes
  # returns:
  #   a tmap-element (plots a map)
  tm_shape(malaria.data)+
    tm_fill(col = variable, 
            breaks = c(0, 500, 1000, 2500, 5000, 10000, 15000, 20000, 25000),
            palette = "YlOrRd",
            title = legend.title)+
    tm_borders(lw = 0.3)+
    tm_layout(legend.position = c(0.75,"bottom"),
              legend.text.size = 0.5,
              legend.title.size = 0.7,
              frame = FALSE)+
    tm_credits(title, 
               position = c(0.2, 0.8), 
               size = 1)
}

# Invoking the function
# 2017 observed and expected malaria cases -------------------------------------
observed_malaria_2017_map <- create.malaria.map(malaria_pop_by_catchment_2017, 
                                                variable = "dr_2017",
                                                title = "2017",
                                                legend.title = "Observed malaria")

expected_malaria_2017_map <- create.malaria.map(expected_malaria_2017,
                                                variable = "expected_2017",
                                                title = "2017",
                                                legend.title = "Expected malaria")

# 2018 observed and expected malaria cases -------------------------------------
observed_malaria_2018_map <- create.malaria.map(malaria_pop_by_catchment_2018,
                                                variable = "dr_2018",
                                                title = "2018",
                                                legend.title = "Observed malaria")

expected_malaria_2018_map <- create.malaria.map(expected_malaria_2018,
                                                variable = "expected_2018",
                                                title = "2018",
                                                legend.title = "Expected malaria")

# 2019 observed and expected malaria cases -------------------------------------
observed_malaria_2019_map <- create.malaria.map(malaria_pop_by_catchment_2019,
                                                variable = "dr_2019",
                                                title = "2019",
                                                legend.title = "Observed malaria")

expected_malaria_2019_map <- create.malaria.map(expected_malaria_2019,
                                                variable = "expected_2019",
                                                title = "2019",
                                                legend.title = "Expected malaria")

# 2020 observed and expected malaria cases -------------------------------------
observed_malaria_2020_map <- create.malaria.map(malaria_pop_by_catchment_2020,
                                                variable = "dr_2020",
                                                title = "2020",
                                                legend.title = "Observed malaria")

expected_malaria_2020_map <- create.malaria.map(expected_malaria_2020,
                                                variable = "expected_2020",
                                                title = "2020",
                                                legend.title = "Expected malaria")

# Layout maps ------------------------------------------------------------------
tmap::tmap_arrange(observed_malaria_2017_map, expected_malaria_2017_map,
                   observed_malaria_2018_map, expected_malaria_2018_map, 
                   observed_malaria_2019_map, expected_malaria_2019_map,
                   observed_malaria_2020_map, expected_malaria_2020_map, ncol = 2)



```

## View SMR by catchment

A ratio greater than 1.0 indicates that more malaria cases have occurred than would have been expected, while a ratio less than 1.0 indicates that less cases have occurred.

```{r, message=FALSE, fig.height = 5, fig.width = 7, fig.cap = 'Fig. 7: Standardised morbidity ratio of malaria by health facility catchment'}

# max(SMR_2017$SMR)
# [1] 7.9
# > max(SMR_2018$SMR)
# [1] 7
# > max(SMR_2019$SMR)
# [1] 7.7
# > max(SMR_2020$SMR)
# [1] 5.3

# Helper function to create maps of SMR by catchment ----------------------------------
create.smr.map <- function(smr.data, 
                           variable = "SMR_category", 
                           title = NA, 
                           legend.title = "SMR"){
  # SMR by catchment map
  # smr.data: sf polygon object containing SMR by catchment data
  # variable: variable name (as character, in qoutes)
  # title: map title in quotes
  # legend.title: legend title in qoutes
  # returns:
  #   a tmap-element (plots a map)
  
  # create category column
  smr.data$SMR_category <- NA
  
  # assigning labels for the SMR estimate legends
  smr.category.list <- c("<0.50", "0.51 to 0.75", "0.76 to 0.99", "1.00", 
                         "1.10 to 1.24", "1.25 to 1.49", ">1.50")
  
  # assigning categories
  smr.data$SMR_category[smr.data$SMR >= 0.0 & smr.data$SMR < 0.49] = -3
  smr.data$SMR_category[smr.data$SMR >= 0.50 & smr.data$SMR < 0.75] = -2
  smr.data$SMR_category[smr.data$SMR >= 0.76 & smr.data$SMR < 0.99] = -1
  smr.data$SMR_category[smr.data$SMR >= 1.00 & smr.data$SMR < 1.09] = 0
  smr.data$SMR_category[smr.data$SMR >= 1.10 & smr.data$SMR < 1.24] = 1
  smr.data$SMR_category[smr.data$SMR >= 1.25 & smr.data$SMR < 1.49] = 2
  smr.data$SMR_category[smr.data$SMR >= 1.50 & smr.data$SMR < 10.0] = 3
  
  # generating divergent colour schemes [Blues - Light Blues – White – Light Reds – Reds]
  # smr.palette <- c("#33a6fe", "#cbe6fe", "#dfeffe", "#feb1b1", "#fe8e8e","#fe0000")
  
  tm_shape(smr.data)+
    tm_fill(col = variable, 
            style = "cat",
            palette = "-RdBu",
            title = legend.title,
            labels = smr.category.list)+
    tm_borders(lw = 0.6)+
    tm_layout(legend.position = c(0.75,"bottom"),
              legend.text.size = 0.5,
              legend.title.size = 0.7,
              frame = FALSE)+
    tm_credits(title, 
               position = c(0.2, 0.8), 
               size = 1)
}

# Invoking function -------------------------------------------------------------------
SMR_2017_map <- create.smr.map(SMR_2017, title = "2017")

SMR_2018_map <- create.smr.map(SMR_2018, title = "2018")

SMR_2019_map <- create.smr.map(SMR_2019, title = "2019")

SMR_2020_map <- create.smr.map(SMR_2020, title = "2020")

# Layout maps -------------------------------------------------------------------------
tmap::tmap_arrange(SMR_2017_map, SMR_2018_map, SMR_2019_map, SMR_2020_map, ncol = 2)


```

## Calculate the proportion of the catchment population living within 1km, 2km, 3km of water bodies

First, using `st_buffer`, we compute 1km, 2km and 3km buffers around dry season water bodies obtained from LandSat satellite imagery using `TropWet` tool in Google Earth Engine. Then geometry of the buffer features are then combined resulting in resolved internal boundaries to enable extracting population values from WorldPop raster. Finally, we calculate the proportion of people in each catchment area living within water bodies.

```{r, message=FALSE, warning=FALSE}
# Combine and transform TropWet derived waterbody polygons -------------------------------
# surface_waterbodies_2017 <- sf::st_as_sf(
#   st_cast(
#     st_union(
#       st_buffer(dryseason_waterbodies_2017, dist = 30)), "POLYGON"))
#
# surface_waterbodies_2018 <- sf::st_as_sf(
#   st_cast(
#     st_union(
#       st_buffer(dryseason_waterbodies_2018, dist = 30)), "POLYGON"))
# 
# surface_waterbodies_2019 <- sf::st_as_sf(
#   st_cast(
#     st_union(
#       st_buffer(dryseason_waterbodies_2019, dist = 30)), "POLYGON"))
# 
# surface_waterbodies_2020 <- sf::st_as_sf(
#   st_cast(
#     st_union(
#       st_buffer(dryseason_waterbodies_2020, dist = 30)), "POLYGON"))

surface_waterbodies_2017 <- sf::st_buffer(dryseason_waterbodies_2017, dist = 30) |>
  sf::st_union() |>
  sf::st_cast("POLYGON") |>
  sf::st_as_sf()

surface_waterbodies_2018 <- sf::st_buffer(dryseason_waterbodies_2018, dist = 30) |>
  sf::st_union() |>
  sf::st_cast("POLYGON") |>
  sf::st_as_sf()

surface_waterbodies_2019 <- sf::st_buffer(dryseason_waterbodies_2019, dist = 30) |>
  sf::st_union() |>
  sf::st_cast("POLYGON") |>
  sf::st_as_sf()

surface_waterbodies_2020 <- sf::st_buffer(dryseason_waterbodies_2020, dist = 30) |>
  sf::st_union() |>
  sf::st_cast("POLYGON") |>
  sf::st_as_sf()


# Helper function to compute 1km, 2km and 3km buffers around the water bodies ---------------------

create.waterbody.buffer <- function(waterbody, distance, catchment){
  # function for creating buffers around waterbodies
  # arguments:
  #   waterbody:  waterbody shapefile
  #   distance: buffer distance in meters
  #   catchment: catchment area shapefile
  # returns:
  #   buffered waterbodies 
  
  # Create buffers around water bodies
  buffer_radius <- sf::st_buffer(waterbody, distance)
  
  # Dissolve the buffers
  # buffer_union <- sf::st_as_sf(st_cast(st_union(buffer_radius),"MULTIPOLYGON"))
  buffer_union <- sf::st_union(buffer_radius) |>
    sf::st_cast("MULTIPOLYGON") |>
    sf::st_as_sf()
  
  # Assign attributes of the 'catchment' to each of the water bodies. 
   buffer_intersect <- sf::st_intersection(buffer_union, catchment)
  
   buffer_intersect_sf <- sf::st_as_sf(buffer_intersect)
   
  # Convert the MULTIPOLYGON object into several POLYGON objects
   # buffer_intersect_polygons <- sf::st_cast(
   #   sf::st_buffer(buffer_intersect_sf,0.0), "MULTIPOLYGON") %>% 
   #   sf::st_cast("POLYGON")
   buffer_intersect_polygons <- sf::st_buffer(buffer_intersect_sf, 0.0) |>
     sf::st_cast("MULTIPOLYGON") |>
     sf::st_cast("POLYGON")
   
  
  # Polygons being seen to be in multiple catchments
   sf::st_intersects(buffer_intersect_polygons, catchment)
  
  # Make the assumption that the attribute is constant throughout the geometry
   sf::st_agr(buffer_intersect_polygons) = "constant"
   
   sf::st_agr(catchment) = "constant"
  
  return(out = buffer_intersect_polygons)
}


# Invoking function
# For 2017 TropWet surface water polygons --------------------------------------------------------
buffer_1km_2017 <- create.waterbody.buffer(waterbody = surface_waterbodies_2017, 
                                           distance = 1000, 
                                           catchment = malire_new)

buffer_2km_2017 <- create.waterbody.buffer(waterbody = surface_waterbodies_2017, 
                                           distance = 2000, 
                                           catchment = malire_new)

buffer_3km_2017 <- create.waterbody.buffer(waterbody = surface_waterbodies_2017, 
                                           distance = 3000,
                                           catchment = malire_new)

# For 2018 TropWet surface water polygons --------------------------------------------------------
buffer_1km_2018 <- create.waterbody.buffer(waterbody = surface_waterbodies_2018, 
                                           distance = 1000, 
                                           catchment = malire_new)

buffer_2km_2018 <- create.waterbody.buffer(waterbody = surface_waterbodies_2018, 
                                           distance = 2000, 
                                           catchment = malire_new)

buffer_3km_2018 <- create.waterbody.buffer(waterbody = surface_waterbodies_2018, 
                                           distance = 3000, 
                                           catchment = malire_new)
 
# For 2019 TropWet surface water polygons ------------------------------------------------------
buffer_1km_2019 <- create.waterbody.buffer(waterbody = surface_waterbodies_2019, 
                                           distance = 1000, 
                                           catchment = malire_new)

buffer_2km_2019 <- create.waterbody.buffer(waterbody = surface_waterbodies_2019, 
                                           distance = 2000, 
                                           catchment = malire_new)

buffer_3km_2019 <- create.waterbody.buffer(waterbody = surface_waterbodies_2019, 
                                           distance = 3000, 
                                           catchment = malire_new)

# For 2020 TropWet surface water polygons ------------------------------------------------------
buffer_1km_2020 <- create.waterbody.buffer(waterbody = surface_waterbodies_2020, 
                                           distance = 1000, 
                                           catchment = malire_new)

buffer_2km_2020 <- create.waterbody.buffer(waterbody = surface_waterbodies_2020, 
                                           distance = 2000, 
                                           catchment = malire_new)

buffer_3km_2020 <- create.waterbody.buffer(waterbody = surface_waterbodies_2020, 
                                           distance = 3000, 
                                           catchment = malire_new)

```

View the created waterbody buffers

```{r, fig.height=9, fig.width=10, fig.cap='Fig 8. Buffers around dry season waterbodies in Kasungu'}

# Map the buffers
create.buffer.map <- function(buffers, boundary = malire_new, title = NA){
  # function for creating buffer map in ggplot
  # arguments:
  #   buffer:  waterbodies buffer polygon layer
  #   boundary: health facility catchment polygons
  #   title: main title
  # returns:
  #   a map-element (plots a map)
  ggplot(data = buffers)+
     geom_sf()+
     geom_sf(data = boundary, 
             fill = NA)+
     theme_void()+
     labs(title = title)
}

# Invoking the function
# For 2017 -------------------------------------------------------------------------------
buffer_1km_2017_map <- create.buffer.map(buffer_1km_2017, title = "2017: 1km Buffers")

buffer_2km_2017_map <- create.buffer.map(buffer_2km_2017, title = "2017: 2km Buffers")

buffer_3km_2017_map <- create.buffer.map(buffer_3km_2017, title = "2017: 3km Buffers")

# For 2018 --------------------------------------------------------------------------------
buffer_1km_2018_map <- create.buffer.map(buffer_1km_2018, title = "2018: 1km Buffers")

buffer_2km_2018_map <- create.buffer.map(buffer_2km_2018, title = "2018: 2km Buffers")

buffer_3km_2018_map <- create.buffer.map(buffer_3km_2018, title = "2018: 3km Buffers")

# For 2019 ---------------------------------------------------------------------------------
buffer_1km_2019_map <- create.buffer.map(buffer_1km_2019, title = "2019: 1km Buffers")

buffer_2km_2019_map <- create.buffer.map(buffer_2km_2019, title = "2019: 2km Buffers")

buffer_3km_2019_map <- create.buffer.map(buffer_3km_2019, title = "2019: 3km Buffers")

# For 2020 --------------------------------------------------------------------------------
buffer_1km_2020_map <- create.buffer.map(buffer_1km_2020, title = "2020: 1km Buffers")

buffer_2km_2020_map <- create.buffer.map(buffer_2km_2020, title = "2020: 2km Buffers")

buffer_3km_2020_map <- create.buffer.map(buffer_3km_2020, title = "2020: 3km Buffers")
 
grid.arrange(buffer_1km_2017_map, buffer_1km_2018_map, buffer_1km_2019_map, buffer_1km_2020_map,
             buffer_2km_2017_map, buffer_2km_2018_map, buffer_2km_2019_map, buffer_2km_2020_map, 
             buffer_3km_2017_map, buffer_3km_2018_map, buffer_3km_2019_map, buffer_3km_2020_map, ncol = 4)


```

Extract the population living within waterbody buffers by catchment area

```{r, message=FALSE, warning=FALSE}

# Helper function to calculate estimated number of people living within waterbody buffers
# in each catchment area
estimate.buffer.pop <- function(catchment.population, buffers, catchment.area){
  
  # Extract population estimates from WorldPop raster
  buffers$buffer_pop <- raster::extract(catchment.population,
                                        buffers, 
                                        fun = sum, 
                                        na.rm = TRUE)
                                               
                                              
  # Find which catchment each polygon belongs to using its centroid - a point dataset 
  # representing the geographic center-points of the polygons 
  # buffer_by_catchment <- st_intersection(st_centroid(buffers), catchment.area)
  buffer_by_catchment <- sf::st_centroid(buffers) |>
    sf::st_intersection(catchment.area)
  
  # Notice that the buffer_catchment is comprised of separate POLYGONS (buffer_by_catchment$x). 
  # The first step is to “dissolve” away these POLYGONS into one MULTIPOLYGON. 
  # There is no sf equivalent to the QGIS or ArcMap “dissolve” operation. 
  # Instead we use a combination of group_by and summarize from the dplyr package. 
  # Stats::aggregate from sf package, and dplyr::summarize both do essentially the same.
   buffer_pop_aggregated <- buffer_by_catchment |> 
     dplyr::group_by(DN) |>
     dplyr::summarize(buffer_pop_aggregated = round(sum(buffer_pop, na.rm = TRUE)))

  buffer_pop <- merge(catchment.area, st_drop_geometry(buffer_pop_aggregated),
                      by = 'DN', all.x = TRUE)
  
  
  
  return(out = buffer_pop)
  
}

# Invoking the function and calculating proportion of 
# catchment population living within buffers
# 2017 buffer population -------------------------------------------------------
buffer_pop_1km_2017 <- estimate.buffer.pop(
  kasungu_population_2017, 
  buffer_1km_2017, 
   malaria_pop_by_catchment_2017) |> 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) |>
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100)) 

buffer_pop_2km_2017 <- estimate.buffer.pop(
  kasungu_population_2017,
  buffer_2km_2017,
  malaria_pop_by_catchment_2017) |>
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) |>
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100)) 

buffer_pop_3km_2017 <- estimate.buffer.pop(
  kasungu_population_2017,
  buffer_3km_2017,
  malaria_pop_by_catchment_2017) |> 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) |>
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

# 2018 buffer population -------------------------------------------------------
buffer_pop_1km_2018 <- estimate.buffer.pop(
  kasungu_population_2018,
  buffer_1km_2018,
  malaria_pop_by_catchment_2018) |> 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) |> 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

buffer_pop_2km_2018 <- estimate.buffer.pop(
  kasungu_population_2018,
  buffer_2km_2018,
  malaria_pop_by_catchment_2018) |> 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) |>
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

buffer_pop_3km_2018 <- estimate.buffer.pop(
  kasungu_population_2018,
  buffer_3km_2018,
  malaria_pop_by_catchment_2018) |> 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) |> 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

# 2019 buffer population -------------------------------------------------------
buffer_pop_1km_2019 <- estimate.buffer.pop(
  kasungu_population_2019,
  buffer_1km_2019,
  malaria_pop_by_catchment_2019) |> 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) |> 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

buffer_pop_2km_2019 <- estimate.buffer.pop(
  kasungu_population_2019,
  buffer_2km_2019,
  malaria_pop_by_catchment_2019) |>
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) |> 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

buffer_pop_3km_2019 <- estimate.buffer.pop(
  kasungu_population_2019,
  buffer_3km_2019,
  malaria_pop_by_catchment_2019) |> 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) |>
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

# 2020 buffer population -------------------------------------------------------
buffer_pop_1km_2020 <- estimate.buffer.pop(
  kasungu_population_2020,
  buffer_1km_2020,
  malaria_pop_by_catchment_2020) |>
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) |> 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

buffer_pop_2km_2020 <- estimate.buffer.pop(
  kasungu_population_2020,
  buffer_2km_2020,
  malaria_pop_by_catchment_2020) |> 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) |> 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

buffer_pop_3km_2020 <- estimate.buffer.pop(
  kasungu_population_2020,
  buffer_3km_2020,
  malaria_pop_by_catchment_2020) |> 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) |> 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

```

Mapping proportion of catchment population living within waterbodies

```{r, message=FALSE, warning = FALSE, fig.height = 9, fig.width = 11, fig.cap='Fig 9. Proportion of catchment population living around water bodies'}

# Helper function to create maps of proportion of people living in proximity ----------
# to water bodies in each catchment area
create.pop.proportion.map <- function(pop.data, 
                                      variable = "prop_buffer_catchment_pop", 
                                      title = NA, 
                                      legend.title = NA){
 
  # pop.data: sf polygon object containing proportion of catchment population 
  #           living within water bodies
  # variable: variable name (as character, in qoutes)
  # title: map title in quotes
  # legend.title: legend title in qoutes
  # returns:
  #   a tmap-element (plots a map)
 
  tm_shape(pop.data)+
    tm_fill(col = variable, 
            breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100),
            palette = "YlOrBr",
            title = legend.title)+
    tm_borders(lw = 0.3)+
    tm_layout(legend.position = c(0.8,"bottom"),
              legend.text.size = 0.5,
              legend.title.size = 0.7,
              frame = FALSE)+
    tm_credits(title, 
               position = c(0.25, 0.75), 
               size = 1)
}

# Invoking function 
# 2017 population proportion ---------------------------------------------------
pop_proportion_1km_2017_map <- create.pop.proportion.map(
  buffer_pop_1km_2017, 
  title = "2017",
  legend.title = "Population within \n1km buffers (%)")

pop_proportion_2km_2017_map <- create.pop.proportion.map(
  buffer_pop_2km_2017, 
  title = "2017",
  legend.title = "Population within \n2km buffers (%)")

pop_proportion_3km_2017_map <- create.pop.proportion.map(
  buffer_pop_3km_2017,
  title = "2017",
  legend.title = "Population within \n3km buffers (%)")

# 2018 population proportion ---------------------------------------------------
pop_proportion_1km_2018_map <- create.pop.proportion.map(
  buffer_pop_1km_2018,
  title = "2018",
  legend.title = "Population within \n1km buffers (%)")

pop_proportion_2km_2018_map <- create.pop.proportion.map(
  buffer_pop_2km_2018,
  title = "2018",
  legend.title = "Population within \n2km buffers (%)")

pop_proportion_3km_2018_map <- create.pop.proportion.map(
  buffer_pop_3km_2018,
  title = "2018",
  legend.title = "Population within \n3km buffers (%)")

# 2019 population proportion ---------------------------------------------------
pop_proportion_1km_2019_map <- create.pop.proportion.map(
  buffer_pop_1km_2019,
  title = "2019",
  legend.title = "Population within \n1km buffers (%)")

pop_proportion_2km_2019_map <- create.pop.proportion.map(
  buffer_pop_2km_2019,
  title = "2019",
  legend.title = "Population within \n2km buffers (%)")

pop_proportion_3km_2019_map <- create.pop.proportion.map(
  buffer_pop_3km_2019,
  title = "2019",
  legend.title = "Population within \n3km buffers (%)")

# 2020 population proportion ---------------------------------------------------
pop_proportion_1km_2020_map <- create.pop.proportion.map(
  buffer_pop_1km_2020,
  title = "2020",
  legend.title = "Population within \n1km buffers (%)")

pop_proportion_2km_2020_map <- create.pop.proportion.map(
  buffer_pop_2km_2020,
  title = "2020",
  legend.title = "Population within \n2km buffers (%)")

pop_proportion_3km_2020_map <- create.pop.proportion.map(
  buffer_pop_3km_2020,
  title = "2020",
  legend.title = "Population within \n3km buffers (%)")

# Layout maps ------------------------------------------------------------------
tmap::tmap_arrange(pop_proportion_1km_2017_map, pop_proportion_2km_2017_map, 
                   pop_proportion_3km_2017_map, pop_proportion_1km_2018_map,
                   pop_proportion_2km_2018_map, pop_proportion_3km_2018_map,
                   pop_proportion_1km_2019_map, pop_proportion_2km_2019_map,
                   pop_proportion_3km_2019_map, pop_proportion_1km_2020_map,
                   pop_proportion_2km_2020_map, pop_proportion_3km_2020_map, ncol = 3)
```

## Scatter plots of SMR against the proportion of the catchment population living waterbody buffers

A correlation coeeficient of more than zero (cor.coeff r \> 0.1) indicates some positive association between the SMR and the buffer population variables. That is, SMR of dry season malaria increases with increase in number of people surrounding water bodies.

```{r, message=FALSE, warning=FALSE, fig.height=15, fig.width=11, fig.cap='Fig 10. Relationship between standardised morbidity ratio and living near waterbodies'}

# Helper function to tidy and bind the SMR and proportion of -------------------
# buffer-catchment population data frames
tidy.data <- function(smr.df, 
                      proportion.pop.1km, 
                      proprotion.pop.2km,
                      proportion.pop.3km){

# Convert the sf objects to data frames-------------------------------------------
smr_df <- as.data.frame(smr.df) |> 
  dplyr::select(rowID, Names, SMR)

proportion_pop_1km_df <- as.data.frame(proportion.pop.1km) |> 
  dplyr::select(rowID, prop_pop_1km = `prop_buffer_catchment_pop`)

proportion_pop_2km_df <- as.data.frame(proprotion.pop.2km) |> 
  dplyr::select(rowID, prop_pop_2km = `prop_buffer_catchment_pop`)

proportion_pop_3km_df <- as.data.frame(proportion.pop.3km) |>
  dplyr::select(rowID, prop_pop_3km = `prop_buffer_catchment_pop`)

# Merge SMR and population data frames -----------------------------------------
combined_1 <- merge(smr_df, proportion_pop_1km_df, by = "rowID", all = TRUE)

combined_2 <- merge(proportion_pop_2km_df, proportion_pop_3km_df)

combined_fully <- merge(combined_1, combined_2, by = "rowID", all = TRUE)

}

# Invoking the function --------------------------------------------------------
smr_pop_2017 <- tidy.data(SMR_2017, buffer_pop_1km_2017, buffer_pop_2km_2017, buffer_pop_3km_2017)

smr_pop_2018 <- tidy.data(SMR_2018, buffer_pop_1km_2018, buffer_pop_2km_2018, buffer_pop_3km_2018)

smr_pop_2019 <- tidy.data(SMR_2019, buffer_pop_1km_2019, buffer_pop_2km_2019, buffer_pop_3km_2019)

smr_pop_2020 <- tidy.data(SMR_2020, buffer_pop_1km_2020, buffer_pop_2km_2020, buffer_pop_3km_2020)

# Helper function to create scatter plots --------------------------------------
create.scatter.plot <- function(smr.pop.df, 
                                independent.var = NA,
                                dependent.var = "SMR",
                                x.label = NA,
                                plot.title = NA){
  
  scatter.plot <- ggpubr::ggscatter(smr.pop.df,          # data frame
                                    x = independent.var, # x-axis variable
                                    y = dependent.var,   # y-axis variable
                                    add = "reg.line",    # Add regression line
                                    conf.int = TRUE,     # Add confidence interval
                                    add.params = list(color = "red",
                                                      fill = "lightgray"),
                                    palette = "jco",     # journal color palette. see ?ggpar
                                    xlab = x.label,      # x-axis label
                                    ylab = "SMR",        # y-axis label
                                    title = plot.title)+    
                  ggpubr::stat_cor(label.y = 4)+         # Add correlation coefficient
                  ggpubr::font("title", size = 10, face = "bold")+
                  ggpubr::font("xlab", size = 10)+
                  ggpubr::font("ylab", size = 10)
 
  return(scatter.plot)
  
}

# Invoking function 
# 2017 scatter plots ------------------------------------------------------------

scatter_1km_2017 <- create.scatter.plot(
  smr_pop_2017, independent.var = "prop_pop_1km",
  x.label = "Percentage of catchment population \nliving in 1km buffer",
  plot.title = "2017")

scatter_2km_2017 <- create.scatter.plot(
  smr_pop_2017, independent.var = "prop_pop_2km",
  x.label = "Percentage of catchment population \nliving in 2km buffer",
  plot.title = "2017")

scatter_3km_2017 <- create.scatter.plot(
  smr_pop_2017, independent.var = "prop_pop_3km",
  x.label = "Percentage of catchment population \nliving in 3km buffer",
  plot.title = "2017")

# 2018 scatter plots -----------------------------------------------------------
scatter_1km_2018 <- create.scatter.plot(
  smr_pop_2018, independent.var = "prop_pop_1km",
  x.label = "Percentage of catchment population \nliving in 1km buffer",
  plot.title = "2018")

scatter_2km_2018 <- create.scatter.plot(
  smr_pop_2018, independent.var = "prop_pop_2km",
  x.label = "Percentage of catchment population \nliving in 2km buffer",
  plot.title = "2018")

scatter_3km_2018 <- create.scatter.plot(
  smr_pop_2018, independent.var = "prop_pop_3km",
  x.label = "Percentage of catchment population \nliving in 3km buffer",
  plot.title = "2018")

# 2019 scatter plots -----------------------------------------------------------
scatter_1km_2019 <- create.scatter.plot(
  smr_pop_2019, independent.var = "prop_pop_1km",
  x.label = "Percentage of catchment population \nliving in 1km buffer",
  plot.title = "2019")

scatter_2km_2019 <- create.scatter.plot(
  smr_pop_2019, independent.var = "prop_pop_2km",
  x.label = "Percentage of catchment population \nliving in 2km buffer",
  plot.title = "2019")

scatter_3km_2019 <- create.scatter.plot(
  smr_pop_2019, independent.var = "prop_pop_3km",
  x.label = "Percentage of catchment population \nliving in 3km buffer",
  plot.title = "2019")

# 2020 scatter plots -----------------------------------------------------------
scatter_1km_2020 <- create.scatter.plot(
  smr_pop_2020, independent.var = "prop_pop_1km",
  x.label = "Percentage of catchment population \nliving in 1km buffer",
  plot.title = "2020")

scatter_2km_2020 <- create.scatter.plot(
  smr_pop_2020, independent.var = "prop_pop_2km",
  x.label = "Percentage of catchment population \nliving in 2km buffer",
  plot.title = "2020")

scatter_3km_2020 <- create.scatter.plot(
  smr_pop_2020, independent.var = "prop_pop_3km",
  x.label = "Percentage of catchment population \nliving in 3km buffer",
  plot.title = "2020")

# Arrange the plots ------------------------------------------------------------
ggpubr::ggarrange(scatter_1km_2017, scatter_2km_2017, scatter_3km_2017,
                  scatter_1km_2018, scatter_2km_2018, scatter_3km_2018,
                  scatter_1km_2019, scatter_2km_2019, scatter_3km_2019, 
                  scatter_1km_2020, scatter_2km_2020, scatter_3km_2020,
                  ncol = 3, nrow = 4)

```

## Model fitting
$$ ln (E(y)) = 𝜷_0 + 𝜷_1 x + ln(𝒆_𝒊)$$
where,
dependent variable, 𝑦 = observed malaria cases;
𝐸(𝑦) = expected count value;
independent variable, 𝑥 = percentage of people living near dams; and
offset, 𝑒= expected malaria cases


Summary outputs:

`Estimate`    : the `intercept` ($𝜷_0$) and the beta coefficient estimates associated to each predictor variable.

`Std.Error`   : the `standard error` of the coefficient estimates. This represents the accuracy of the coefficients. The larger the standard error, the less confident we are about the estimate.

`t value`     : the `t-statistic`, which is the coefficient estimate (column 2) divided by the standard error of the estimate (column 3). For a given the predictor, the `t-statistic` evaluates whether or not there is significant association between the predictor and the outcome variable, i.e., whether the beta coefficient of the predictor is significantly different from zero.

`Pr(>|t|)`    : The `p-value` corresponding to the `t-statistic`. The smaller the `p-value`, the more significant the estimate is.

`Residuals`   : Provide a quick view of the distribution of the residuals, which by definition have a mean zero. Therefore, the `median` should not be far from zero, and the minimum (`min`)and maximum (`max`) should be roughly equal in absolute value.

`Coefficients`: Shows the regression beta coefficients and their statistical significance. Predictor variables, that are significantly associated to the outcome variable, are marked by stars.

`Residual standard error` (`RSE`), and `R-squared` ($R^2$)  metrics tell how well the model fits to our data. An (adjusted) $R^2$ that is close to 1 indicates that a large proportion of the variability in the outcome has been explained by the regression model. A number near 0 indicates that the regression model did not explain much of the variability in the outcome.


```{r, message=FALSE}
# Combine data for model fitting -----------------------------------------------
model_data_2017 <- merge(expected_malaria_2017, smr_pop_2017, by = "rowID", all = TRUE) |>
  dplyr::select(-Names.y) |> 
  dplyr::rename(Names = Names.x)

model_data_2018 <-  merge(expected_malaria_2018, smr_pop_2018, by = "rowID", all = TRUE) |> 
  dplyr::select(-Names.y) |> 
  dplyr::rename(Names = Names.x)

model_data_2019 <-  merge(expected_malaria_2019, smr_pop_2019, by = "rowID", all = TRUE) |>
  dplyr::select(-Names.y) |> 
  dplyr::rename(Names = Names.x)

model_data_2020 <-  merge(expected_malaria_2020, smr_pop_2020, by = "rowID", all = TRUE) |> 
  dplyr::select(-Names.y) |>
  dplyr::rename(Names = Names.x)

# Normality test ---------------------------------------------------------------
# Check whether the dependent variable follows a poisson distribution
# Dry season malaria cases appear to be non normally distributed i.e highly skewed 

graphics::hist(model_data_2017$observed_2017, 
               main = "Histogram of 2017 dry season malaria cases ",
               xlab = "Malaria cases",
               las = 1)

graphics::hist(model_data_2018$observed_2018)

graphics::hist(model_data_2019$observed_2019)

graphics::hist(model_data_2020$observed_2020)

# Fit generalised linear model -------------------------------------------------
# Defining model parameters:
# response variable: observed_2017, observed_2018, observed_2019, observed_2020 are
#                    recorded dry season malaria cases in that year
# risk factor: prop_pop_1km,  prop_pop_2km,  prop_pop_3km are the percentage of people living
#             within 1km, 2km and 3km buffers of water bodies, respectively.
# offset: expected_* is the number of malaria cases we would expect if the malaria rate
#          was equal in all the catchment areas

# 2017 -------------------------------------------------------------------------
model_1km_2017 <- glm(observed_2017~1+prop_pop_1km+offset(log(expected_2017)),
                      data = model_data_2017, family = 'poisson')

# Display the statistical summary of the model
summary(model_1km_2017)
 
# The estimated regression equation can be written as follow: 
# observed malaria cases = -0.2223780 + 0.0247164 * percentage of catchment population living near dams. 
# Using this formula, for each change in number of people living near dams,
# we can predict the number of dry season malaria case.
#
# For example:
#   for a buffer population equal zero, we can expect -0.22 malaria cases. 
#   for a buffer population equal 1000, we can expect  round(-0.22 + 0.02 * 1000) = 20 malaria cases.

sjPlot::tab_model(model_1km_2017, digits = 3, digits.re = 3)

# From the output above, the R2 is 1, meaning that the observed and the predicted 
# outcome values are highly correlated, which is very good. 

model_2km_2017 <- glm(observed_2017~1+prop_pop_2km+offset(log(expected_2017)),
                      data = model_data_2017, family = 'poisson')

summary(model_2km_2017)
sjPlot::tab_model(model_2km_2017, digits = 3, digits.re = 3)

model_3km_2017 <- glm(observed_2017~1+prop_pop_3km+offset(log(expected_2017)),
                      data = model_data_2017, family = 'poisson')

summary(model_3km_2017)
sjPlot::tab_model(model_3km_2017, digits = 3, digits.re = 3)

# 2018 -------------------------------------------------------------------------
model_1km_2018 <- glm(observed_2018~1+prop_pop_1km+offset(log(expected_2018)),
                      data = model_data_2018, family = 'poisson')

summary(model_1km_2018)
sjPlot::tab_model(model_1km_2018, digits = 3, digits.re = 3)

model_2km_2018 <- glm(observed_2018~1+prop_pop_2km+offset(log(expected_2018)),
                      data = model_data_2018, family = 'poisson')

summary(model_2km_2018)
sjPlot::tab_model(model_2km_2018, digits = 3, digits.re = 3)

model_3km_2018 <- glm(observed_2018~1+prop_pop_3km+offset(log(expected_2018)),
                      data = model_data_2018, family = 'poisson')

summary(model_3km_2018)
sjPlot::tab_model(model_3km_2018, digits = 3, digits.re = 3)

# 2019 -------------------------------------------------------------------------
model_1km_2019 <- glm(observed_2019~1+prop_pop_1km+offset(log(expected_2019)),
                      data = model_data_2019, family = 'poisson')

summary(model_1km_2019)
sjPlot::tab_model(model_1km_2019, digits = 3, digits.re = 3)


model_2km_2019 <- glm(observed_2019~1+prop_pop_2km+offset(log(expected_2019)),
                      data = model_data_2019, family = 'poisson')

summary(model_2km_2019)
sjPlot::tab_model(model_2km_2019, digits = 3, digits.re = 3)

model_3km_2019 <- glm(observed_2019~1+prop_pop_3km+offset(log(expected_2019)),
                      data = model_data_2019, family = 'poisson')

summary(model_3km_2019)
sjPlot::tab_model(model_3km_2019, digits = 3, digits.re = 3)

# 2020 -------------------------------------------------------------------------
model_1km_2020 <- glm(observed_2020~1+prop_pop_1km+offset(log(expected_2020)),
                      data = model_data_2020, family = 'poisson')

summary(model_1km_2020)
sjPlot::tab_model(model_1km_2020, digits = 3, digits.re = 3)

model_2km_2020 <- glm(observed_2020~1+prop_pop_2km+offset(log(expected_2020)),
                      data = model_data_2020, family = 'poisson')

summary(model_2km_2020)
sjPlot::tab_model(model_2km_2020, digits = 3, digits.re = 3)

model_3km_2020 <- glm(observed_2020~1+prop_pop_3km+offset(log(expected_2020)),
                      data = model_data_2020, family = 'poisson')

summary(model_3km_2020)
sjPlot::tab_model(model_3km_2020, digits = 3, digits.re = 3)

```

## Assess the performance and accuracy of the regression model. 
Here, we evaluate how well the `Poisson regression` model is predicting the outcome of a new test data that have not been used to build the model i.e how close the prediction is close to the real value.

Two important metrics have been used to assess the performance of the predictive regression model:

`Root Mean Squared Error`, which measures the model prediction error. It corresponds to the average difference between the observed known values of the outcome and the predicted value by the model.`RMSE` is computed as `RMSE = mean((observed - predicted)^2) %>% sqrt()`. The lower the `RMSE`, the better the model.


$$RMSE = \sqrt \frac {\Sigma ( y - \hat{y} )^2}{N}$$



`R-square`, representing the squared correlation between the observed known outcome values and the predicted values by the model. The higher the $R^2$, the better the model. 


$$R^2 = \frac {\Sigma {(y - \bar y)^2} - \Sigma {(y - \hat y)^2}}{\Sigma {(y - \bar y)^2}}$$

```{r, message=FALSE, warning=FALSE, fig.height=17}

# Tidy model assessment data -----------------------------------------
model_assessment_data_2017 <- model_data_2017 |>
  dplyr::as_tibble() |>
  dplyr::select(Names, observed_2017, expected_2017,
                prop_pop_1km, prop_pop_2km, prop_pop_3km) |>
  dplyr::glimpse()

model_assessment_data_2018 <- model_data_2018 |>
  dplyr::as_tibble() |>
  dplyr::select(Names, observed_2018, expected_2018,
                prop_pop_1km, prop_pop_2km, prop_pop_3km) |>
  dplyr::glimpse()


model_assessment_data_2019 <- model_data_2019 |>
  dplyr::as_tibble() |>
  dplyr::select(Names, observed_2019, expected_2019,
                prop_pop_1km, prop_pop_2km, prop_pop_3km) |>
  dplyr::glimpse()

model_assessment_data_2020 <- model_data_2020 |>
  dplyr::as_tibble() |>
  dplyr::select(Names, observed_2020, expected_2020,
                prop_pop_1km, prop_pop_2km, prop_pop_3km) |>
  dplyr::glimpse()

# Split the data into training and test set 
# 80% of the data is used for training
set.seed(123) # generate a sequence of random numbers
# 2017 -------------------------------------------------------------------------
training_samples_2017 <- model_assessment_data_2017$observed_2017 |>
  caret::createDataPartition(p = 0.8, list = FALSE)

train_data_2017  <- model_assessment_data_2017[training_samples_2017, ]

test_data_2017 <- model_assessment_data_2017[-training_samples_2017, ]

# 2018 
training_samples_2018 <- model_assessment_data_2018$observed_2018 |>
  caret::createDataPartition(p = 0.8, list = FALSE) # see ?createDataPartition

train_data_2018 <- model_assessment_data_2018[training_samples_2018, ]

test_data_2018 <- model_assessment_data_2018[-training_samples_2018, ]

# 2019 
training_samples_2019 <- model_assessment_data_2019$observed_2019 |>
  caret::createDataPartition(p = 0.8, list = FALSE)

train_data_2019 <- model_assessment_data_2019[training_samples_2019, ]

test_data_2019 <- model_assessment_data_2019[-training_samples_2019, ]

# 2020 
training_samples_2020 <- model_assessment_data_2020$observed_2020 |>
  caret::createDataPartition(p = 0.8, list = FALSE)

train_data_2020 <- model_assessment_data_2020[training_samples_2020, ]

test_data_2020 <- model_assessment_data_2020[-training_samples_2020, ]

# Make predictions using the test data in order to evaluate the performance 
# of our regression model aka goodness-of-fit
# 2017 -------------------------------------------------------------------------
predictions_1km_2017 <- model_1km_2017 |>
  stats::predict(test_data_2017)

predictions_2km_2017 <- model_2km_2017 |>
  stats::predict(test_data_2017)

# 2018 
predictions_1km_2018 <- model_1km_2018 |>
  stats::predict(test_data_2018)

predictions_2km_2018 <- model_2km_2018 |>
  stats::predict(test_data_2018)

# 2019 
predictions_1km_2019 <- model_1km_2019 |>
  stats::predict(test_data_2019)

predictions_2km_2019 <- model_2km_2019 |>
  stats::predict(test_data_2019)

# 2020 
predictions_1km_2020 <- model_1km_2020 |>
  stats::predict(test_data_2020)

predictions_2km_2020 <- model_2km_2020 |>
  stats::predict(test_data_2020)

# Model performance ------------------------------------------------------------
# (a) Compute the prediction error, RMSE. The lower the RMSE, the better the model
# 2017
caret::RMSE(predictions_1km_2017, test_data_2017$observed_2017)

caret::RMSE(predictions_2km_2017, test_data_2017$observed_2017)

# 2018 
caret::RMSE(predictions_1km_2018, test_data_2018$observed_2018)

caret::RMSE(predictions_2km_2018, test_data_2018$observed_2018)

# 2019 
caret::RMSE(predictions_1km_2019, test_data_2019$observed_2019)

caret::RMSE(predictions_2km_2019, test_data_2019$observed_2019)

# 2020 
caret::RMSE(predictions_1km_2020, test_data_2020$observed_2020)

caret::RMSE(predictions_2km_2020, test_data_2020$observed_2020)

# (b) Compute R-square ---------------------------------------------------------
# 2017 
caret::R2(predictions_1km_2017, test_data_2017$observed_2017)
# [1] 0.8549051
# From the output above, the R2 is 0.85, meaning that the observed and the 
# predicted outcome values are highly correlated

caret::R2(predictions_2km_2017, test_data_2017$observed_2017)

# 2018 
caret::R2(predictions_1km_2018, test_data_2018$observed_2018)
# [1] 0.4304403
# From the output above, the R2 is 0.43, meaning that the observed and the 
# predicted outcome values are less correlated

caret::R2(predictions_2km_2018, test_data_2018$observed_2018)

# 2019 
caret::R2(predictions_1km_2019, test_data_2019$observed_2019)

caret::R2(predictions_2km_2019, test_data_2019$observed_2019)

# 2020 
caret::R2(predictions_1km_2020, test_data_2020$observed_2020)

caret::R2(predictions_2km_2020, test_data_2020$observed_2020)

# Check for collinearity, normality or heteroscedasticity() --------------------

performance::check_model(model_1km_2017)

performance::check_model(model_2km_2017)

performance::check_model(model_1km_2018)

performance::check_model(model_2km_2018)

performance::check_model(model_1km_2019)

performance::check_model(model_2km_2019)

performance::check_model(model_1km_2020)

performance::check_model(model_2km_2020)

# Model performance summaries --------------------------------------------------
# Compute indices of model performance for the regression models and 
# compare the quality of the models.
# Note that all score value do not necessarily sum up to 100%. See ?compare_performance
model_comparison <- performance::compare_performance(model_1km_2017, model_2km_2017,
                                                     model_1km_2018, model_2km_2018, 
                                                     model_1km_2019, model_2km_2019,
                                                     model_1km_2020, model_2km_2020, 
                                                     metrics = "common", rank = TRUE)

model_comparison

```

Visualization of indices of models’ performance

```{r, message=FALSE, warning=FALSE, fig.width=11}
plot(performance::compare_performance(model_1km_2017, model_2km_2017,
                                      model_1km_2018, model_2km_2018, 
                                      model_1km_2019, model_2km_2019,
                                      model_1km_2020, model_2km_2020, 
                                      metrics = "common", rank = TRUE))
```


Check how well the fitted values line up with the observations

The fitted values appear to line up particularly well with the observed data, suggesting that `prop_pop_*` (i.e., proportion of catchment population living near water bodies) can help us understand malaria risk in the catchment areas.

```{r, message=FALSE, fig.height=15, fig.width=11, fig.cap='Fig 11. How well the percentage of catchment population living around water bodies explain observed malaria incidence'}
# Helper function to create scatter plots to see how well 
# fitted values line up with observed malaria cases
plot.fitted.values <- function(fitted.values.df, model.df, title){
  
  # Remove missing values from model data since 
  # model fitting deletes missing observations
  model.df.complete <- model.df %>% 
    tidyr::drop_na() %>% 
    dplyr::rename_at(vars(starts_with("observed_")), ~ str_c("observed"))

  # Plot fitted versus observed values
  scatter.plot <- ggplot2::ggplot()+ 
                  ggplot2::geom_point(aes(fitted.values.df$fitted.values, 
                                      model.df.complete$observed))+
                  ggplot2::theme_classic()+
                  ggplot2::labs(y = "Observed values", 
                                x = "Fitted values",
                                title = title)
  return(scatter.plot)
}

# Invoking function 
# 2017 -------------------------------------------------------------------------
fitted_1km_2017 <- plot.fitted.values(model_1km_2017, model_data_2017, "2017: 1km model")

fitted_2km_2017 <- plot.fitted.values(model_2km_2017, model_data_2017, "2017: 2km model")

fitted_3km_2017 <- plot.fitted.values(model_3km_2017, model_data_2017, "2017: 3km model")

# 2018 -------------------------------------------------------------------------
fitted_1km_2018 <- plot.fitted.values(model_1km_2018, model_data_2018, "2018: 1km model")

fitted_2km_2018 <- plot.fitted.values(model_2km_2018, model_data_2018, "2018: 2km model")

fitted_3km_2018 <- plot.fitted.values(model_3km_2018, model_data_2018, "2018: 3km model")

# 2019 -------------------------------------------------------------------------
fitted_1km_2019 <- plot.fitted.values(model_1km_2019, model_data_2019, "2019: 1km model")

fitted_2km_2019 <- plot.fitted.values(model_2km_2019, model_data_2019, "2019: 2km model")

fitted_3km_2019 <- plot.fitted.values(model_3km_2019, model_data_2020, "2019: 3km model")

# 2020 -------------------------------------------------------------------------
fitted_1km_2020 <- plot.fitted.values(model_1km_2020, model_data_2020, "2020: 1km model")

fitted_2km_2020 <- plot.fitted.values(model_2km_2020, model_data_2020, "2020: 2km model")

fitted_3km_2020 <- plot.fitted.values(model_3km_2020, model_data_2020, "2020: 3km model")

# Layout scatter plots ---------------------------------------------------------
cowplot::plot_grid(fitted_1km_2017, fitted_1km_2018, fitted_2km_2018, fitted_3km_2018,
                   fitted_1km_2019, fitted_2km_2019, fitted_1km_2020, ncol = 3, nrow = 3)

```

## Test for residual spatial autocorrelation using adjacency as criterion

```{r, fig.keep='all', fig.cap='Fig 12. Neighbourhood matrix'}
# Find adjacent polygons,
# Contiguity neighbors - all that share a boundary point
catchment_neighbours <- spdep::poly2nb(model_data_2017)  # Queen contiguity

summary(catchment_neighbours)

# Get coordinates from catchment polygons
catchment_points <- as(model_data_2017, "Spatial")

coords <- coordinates(catchment_points)

# Get catchment boundaries and convert to spatial object
catchment_polygons <- as(model_data_2017, "Spatial")

# View the neighbors
{plot(catchment_polygons, asp = 1)+
plot(catchment_neighbours, coords, col = "blue", add = TRUE)}

# Run a Conditional Autoregressive (CAR) model, which allows us to incorporate 
# the spatial autocorrelation between neighbours within our GLM

# First, generate a weights matrix from a neighbours list with spatial weights
adj_matrix <- spdep::nb2mat(catchment_neighbours, style = "B") # see ?nb2mat

# Match row and column names with those of geographic location index 
rownames(adj_matrix) <- colnames(adj_matrix) <- model_data_2017$rowID
# row.names(adj_matrix) <- NULL # alternatively

# Now we can fit the model. The spatial effect is called using the adjacency function which 
# requires the grouping factor (i.e. the rowID of each catchment area)

CAR_model_1km_2017 <- spaMM::fitme(observed_2017~prop_pop_1km+offset(log(expected_2017)),
                                   adjMatrix = adj_matrix, 
                                   data = model_data_2017, 
                                   family = 'poisson')

# Generate 95% CI
coefs <- as.data.frame(summary(CAR_model_1km_2017)$beta_table)


# Moran's I contiguity test
MI_2017 <- spdep::moran(model_data_2017$observed_2017, nb2listw(catchment_neighbours),
                        length(model_data_2017$observed_2017),Szero(nb2listw(catchment_neighbours)))



```


## Yearly variation in malaria cases as a risk factor

```{r, message=FALSE, eval=FALSE}
# Gather year ------------------------------------------------------------------
model_data_2017_longer <- as.data.frame(model_data_2017) %>%
  dplyr::select(rowID, Names, observed_2017, dr_2018, dr_2019, dr_2020, 
                expected_2017, prop_pop_1km, prop_pop_2km, prop_pop_3km) %>% 
  dplyr::mutate(observed = observed_2017) %>% 
  dplyr::rename(`2017` = observed_2017,
                `2018` = dr_2018,
                `2019` = dr_2019,
                `2020` = dr_2020) %>%
  as.data.frame() %>% 
  tidyr::pivot_longer(cols = `2017`:`2020`,
                      names_to = 'year',
                      values_to = 'malaria_cases')

# Model fitting ----------------------------------------------------------------
multivariate_1km_2017 <- glm(observed~1+prop_pop_1km+year+offset(log(expected_2017)),
                             data = model_data_2017_longer, family = 'poisson')

summary(multivariate_1km_2017)
sjPlot::tab_model(multivariate_1km_2017)


multivariate_2km_2017 <- glm(observed~1+prop_pop_2km+year+offset(log(expected_2017)),
                             data = model_data_2017_longer, family = 'poisson')

summary(multivariate_2km_2017)
sjPlot::tab_model(multivariate_2km_2017)

multivariate_3km_2017 <- glm(observed~1+prop_pop_3km+year+offset(log(expected_2017)),
                             data = model_data_2017_longer, family = 'poisson')

summary(multivariate_3km_2017)
sjPlot::tab_model(multivariate_3km_2017)

```
