---
title: "Analysis of Role of Dams on Recorded Dry Season Malaria Incidences in Kasungu"
output: html_document
latex_engine: MiKTeX
author: Clinton Nkolokosa
date: " Last edited `r format(Sys.time(),'%d %B %Y')`"
---
This geospatial statistical model uses routinely collected malaria case data, population data and remotely sensed data, such as open and vegetated water bodies, to estimate population living around open water bodies, expected malaria cases, and `standardised morbidity ratio` (SMR) of malaria. And ultimately, quantify the association between proximity to larval habitat and malaria risk in health facility catchment areas in Kasungu. The SMR compares the risk of morbidity in a population of interest with that of a standard population. In this case, our interest is to find out whether the number of dry season malaria cases in each catchment area are greater than we would expect given the malaria rate for the entire Kasungu district.

We do this by comparing what we *observe* (O) with what we would *expect* (E) if the risk of malaria was equal throughout Kasungu. The SMR statistical notation of catchment *i* can be written as follows:
$$SMR_i = \frac{O_i}{E_i}$$

Buffers around waterbodies are created and then combined with population data in raster format to estimate the proprtion of catcment population living within 1km, 2km and 3km of water bodies. Subsequently, the observed malaria cases are modeled using `Poisson regression` to find out if living within various distances from water bodies is causing variability in malaria risk in Kasungu district. We hypothesize that the risk of being a case in a catchment is dependent on proximity to water bodies. The data used spans from 2017 to 2020 and was derived from digitized DHIS2 malaria records, accessibility mapping, aggregated population geospatial layer and TropWet tool in Google Earth Engine. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', echo = TRUE)
```

### Load packages

Loading the R packages that will be used to read in, view, transform and model the malaria cases and spatial datasets.

```{r, warning = FALSE, message = FALSE}
library(spatialEco)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(plotly)
library(lubridate)
library(knitr)
library(raster)
library(rgdal)
library(rgeos)
library(sf)
library(sp)
library(tmap)
library(spdep)
library(maptools)
library(gridExtra)
library(grid)
library(exactextractr)
library(DataExplorer)
library(mapview)
`%>%` <- magrittr::`%>%`
```

### Tell R where the data is

```{r}
here::here()
```

### Load datasets

The total dry season malaria cases recorded at health-care facilities in Kasungu from 2017 to 2019 are contained in the `KasunguData.csv` sourced from https://dhis2.health.gov.mw/. The `kasungu_facility_catchments_2004.shp` shapefile also contains the population and health information within each health-facility catchment area in Kasungu district. 

The aggregated population raster layers for Malawi e.g.,`ku_pop_2017_1km_aggregated.tif` were downloaded from the Open Spatial and Demographic and Data Research website: https://www.worldpop.org/geodata/country?iso3=MWI. These layers estimate total number of people per grid-cell. The units are number of people per pixel with country totals adjusted to match the corresponding official United Nations population estimates. The datasets were downloaded in Geotiff at a resolution of 1km and are projected in Geographic Coordinate System, WGS84. 

The `kasungu_water.shp`and `water_bodies` layers contain open and vegetated waterbodies polygons, detected using the Tropical Wetland Unmixing Tool (TropWet). TropWet is a Google Earth Engine hosted toolbox that uses the Landsat archive to map tropical wetlands and can be accessed through: https://www.aber.ac.uk/en/dges/research/earth-observation-laboratory/research/tropwet/

```{r, Kasungu water bodies polygons, malaria cases by health facility catchments and population layers, message= FALSE}

# 2017, 2018 and 2019 dry season malaria cases 
dry_season_malaria_2015_2019 <- read.csv(here::here("data", "dry_season_malaria 2015-2019.csv"))

# Export 2017, 2018 and 2019 dry season malaria cases
write.csv(dry_season_malaria_2015_2019, file = "data/dry_season_malaria_2017_2019.csv")

write.table(dry_season_malaria_2015_2019, 
            file = "dry_season_malaria_2017_2019.txt", 
            sep = ",", 
            quote = FALSE, 
            row.names = FALSE)

# 2020 dry season malaria cases
ku_malaria_2020 <- read.csv(here::here("data", "dry_season_malaria_2020.csv"))


# Merge 2015 to 2019 dry season malaria case data with 2020 data
dry_season_malaria_2017_2020 <- cbind.data.frame(dry_season_malaria_2015_2019, ku_malaria_2020) 
 
dry_season_malaria_2017_2020 <- dry_season_malaria_2017_2020[,c("FID", "Names", "dr_2017", 
                                                                "dr_2018", "dr_2019", "dr_2020",
                                                                "LONGITU", "LATITUD")]

# Export 'dry season malaria 2017-2020' as csv
write.csv(dry_season_malaria_2017_2020, file = "data/dry_season_malaria_2017_2019.csv")


# Kasungu district boundary shapefile 
kasungu_district <- st_read(here::here("data", "kasungu_district.shp"))
  
# Kasungu health facility catchments generated from accessibility mapping

malire_new <- sf::st_read(here::here("data", "new_catchments.shp")) %>% 
              sf::st_transform(32736) # reproject to WGS UTM Zone 36 South

# Kasungu population raster layer
kasungu_population_2017 <- raster(here::here("data", "ku_pop_2017_1km_aggregated.tif"))

kasungu_population_2018 <- raster(here::here("data", "ku_pop_2018_1km_aggregated.tif"))

kasungu_population_2019 <- raster(here::here("data", "ku_pop_2019_1km_aggregated.tif"))

kasungu_population_2020 <- raster(here::here("data", "ku_pop_2020_1km_aggregated.tif"))

# Read in waterbodies polygons 
dryseason_waterbodies_2017 <- st_read(here::here("data", "water_bodies_2017.shp"))

dryseason_waterbodies_2018 <- st_read(here::here("data", "kasungu_2018_water.shp"))

dryseason_waterbodies_2019 <- st_read(here::here("data", "kasungu_2019_water.shp"))

dryseason_waterbodies_2020 <- st_read(here::here("data", "water_bodies_2020.shp"))

# Add a field ID to water bodies polygons 
dryseason_waterbodies_2017$ID <- 1:nrow(dryseason_waterbodies_2017)

dryseason_waterbodies_2018$ID <- 1:nrow(dryseason_waterbodies_2018)

dryseason_waterbodies_2019$ID <- 1:nrow(dryseason_waterbodies_2019)

dryseason_waterbodies_2020$ID <- 1:nrow(dryseason_waterbodies_2020)

```

### View the dry season malaria case data

We observe that Kasungu district has 30 health facilities classified as dispensary, health centre, district hospital and rural hospital, and the highest malaria cases were recorded at Kasungu District Hospital. 

```{r, warning = FALSE, message = FALSE, fig.height = 10, fig.width = 8, fig.cap = 'Fig.1 The total malaria cases recorded at each health-care facility in Kasungu district'}

dry_season_malaria_2017_2020 %>% 
  summary()

dry_season_malaria_2017_2020 %>%  
  plotly::plot_ly(y = ~Names,
                  x = ~dr_2017,
                  type = "bar",
                  orientation = 'h',
                  name = "2017") %>%
  plotly::add_trace(x = ~ dr_2018,
                    name = "2018") %>%
  plotly::add_trace(x = ~ dr_2019,
                    name = "2019") %>% 
  plotly::add_trace(x = ~ dr_2020,
                    name = "2020") %>% 
  plotly::layout(#barmode = "stack",
                 xaxis = list(title = "Total malaria cases"),
                 yaxis = list(title = ""),
                 hovermode = "compare",
                 margin = list(b = 10,
                               t = 10,
                               pad = 2))

```


### Kasungu health-care facilities and their catchment areas
Heath facility catchment area is the area from which a health facility attracts patients. The new health facility catchments polygon was generated from generic accessibility mapping script adapted from https://malariaatlas.org/wp-content/uploads/accessibility/R_generic_accessibilty_mapping_script.r The script requires two user supplied datasets: the `2015 friction surface`, which is available here:  http://www.map.ox.ac.uk/accessibility_to_cities/, and a user-supplied .csv of points `dry_season_malaria_2017_2020`. The accumulated cost algorithm `accCost` and `r.Cost` algorithm in GRASS GIS were run to make the final output map of new health facility catchment boundaries.
```{r, warning = FALSE, message = FALSE}

# Using the complete.cases() function to select health centres with complete 
# longitude and latitude coordinates.

zipatala_aggregated <- dry_season_malaria_2017_2020[complete.cases(dry_season_malaria_2017_2020),] %>% 
  dplyr::filter(Names != "Kasalika Health Centre",
                Names != "Bua Health Centre",
                Names != "Kaluluma Rural Hospital") 

# The health facilities close to each other e.g., Kasalika Health Centre and 
# Kasungu District Hospital, and Kaluluma Rural Hospital and Nkhamenya Rural Hospital
# were aggregated in order to generate catchment areas that are geographically correct

zipatala_aggregated$dr_2017[which(zipatala_aggregated$Names == "Kasungu District Hospital")] <- 4528 + 16289

zipatala_aggregated$dr_2018[which(zipatala_aggregated$Names == "Kasungu District Hospital")] <- 4493 +15821

zipatala_aggregated$dr_2019[which(zipatala_aggregated$Names == "Kasungu District Hospital")] <- 2729 + 10721

zipatala_aggregated$dr_2020[which(zipatala_aggregated$Names == "Kasungu District Hospital")] <- 4368 + 24424
  
zipatala_aggregated$dr_2017[which(zipatala_aggregated$Names == "Nkhamenya Rural Hospital")] <- 2887 + 752

zipatala_aggregated$dr_2018[which(zipatala_aggregated$Names == "Nkhamenya Rural Hospital")] <- 851 + 3689

zipatala_aggregated$dr_2019[which(zipatala_aggregated$Names == "Nkhamenya Rural Hospital")] <- 533 + 4004

zipatala_aggregated$dr_2020[which(zipatala_aggregated$Names == "Nkhamenya Rural Hospital")] <- 3587 + 5929

zipatala_aggregated$dr_2017[which(zipatala_aggregated$Names == "Mziza Health Centre")] <- 3863 + 3489

zipatala_aggregated$dr_2018[which(zipatala_aggregated$Names == "Mziza Health Centre")] <- 2815 + 1804

zipatala_aggregated$dr_2019[which(zipatala_aggregated$Names == "Mziza Health Centre")] <- 2439 + 1740

zipatala_aggregated$dr_2020[which(zipatala_aggregated$Names == "Mziza Health Centre")] <- 6194 + 2397

# write.csv(zipatala_aggregated, "data/health_facilities_aggregated.csv")


health_facility_aggr_sf <- sf::st_as_sf(zipatala_aggregated,
                                        coords = c("LONGITU", "LATITUD"),
                                        crs = 4326, agr = "constant")

# st_write(health_facility_aggr_sf, "data/health_facilities_aggregated.shp")

```


View location of the health facilities in the new catchment areas

```{r, message=FALSE, fig.cap='Fig 2. Kasungu health-care facilities and catchment areas'}

# Plot map
tm_shape(malire_new)+
  tm_polygons()+
  tm_shape(health_facility_aggr_sf)+
  tm_dots(size = .3, 
          col = "blue", 
          alpha = 0.5)+
  tm_text("Names", 
          size = .3, 
          just = "top", 
          col = "black", 
          remove.overlap = TRUE)+
  tm_layout(frame = FALSE,
            title = "New Kasungu health facility \n catchment boundaries",
            title.size = .8, 
            title.position = c("left", "top"))+
  tm_compass(position=c("right", "top"))+
  tm_scale_bar(breaks = c(0, 10, 20), 
               text.size = .5)

```


### Kasungu district estimated population per grid-cell

```{r, message = FALSE, warning = FALSE, fig.height = 9, fig.width = 10, fig.cap = 'Fig.3 Estimated total number of people per 1km grid-cell'}

# Take a glimpse at the WorldPop raster layers
kasungu_population_2017

kasungu_population_2018

kasungu_population_2019

kasungu_population_2020

# Helper function to create a raster population map
create.population.map <- function(population.raster, title){
  # raster population map
  # arguments:
  #   population.raster:  aggregated population raster layer from WorldPop
  #   legend.title: legend title
  # returns:
  #   a tmap-element (plots a map)
  tm_shape(population.raster)+
    tm_raster(palette = "YlOrBr", 
              title = title,
              breaks = c(0,100,200,400,600,800,1000,2000,4000,6000,8000))+
    tm_layout(legend.position = c("right", "bottom"),
              frame = FALSE)+
    tm_scale_bar(position = c("left", "bottom"))
}
# Set to static map
tmap_mode("plot")

estimated_pop_2017 <- create.population.map(kasungu_population_2017, title = "2017 Population")

estimated_pop_2018 <- create.population.map(kasungu_population_2018, title = "2018 Population")

estimated_pop_2019 <- create.population.map(kasungu_population_2019, title = "2019 Population")

estimated_pop_2020 <- create.population.map(kasungu_population_2020, title = "2020 Population")

# Layout the maps
tmap_arrange(estimated_pop_2017, estimated_pop_2018, estimated_pop_2019, estimated_pop_2020, nrow = 2) 

```


### Assign dry season malaria cases and population density to new health facility catchments

The WorldPop `aggregated population` e.g. `kasungu_population_2017.tif`, and DHIS2 malaria `dry_season_malaria_2017_2020`  datasets are assigned to the new health facility catchments. 

```{r, message=FALSE, warning=FALSE}
# Helper function that assigns malaria data from health facilities to their catchments areas ----------------
assign.malaria.data <- function(catchment_boundary, malaria_data){
  # arguments:
  #   catchment_boundary: sf polygon object of new catchment boundaries
  #   malaria_data: sf point object with a data frame containing the dry season malaria cases
  # returns:
  #   catchments_malaria_sf: sf polygon object with a data frame containing dry season malaria cases


  # Convert sf objects to spatial
  catchment_shp <- as(catchment_boundary, "Spatial")
  
  malaria_shp <- as(malaria_data, "Spatial")

  # Match CRS
  malaria_shp <- spTransform(malaria_shp, crs(catchment_shp))

  # Overlay aggregated health facility points and extract 2017 - 2020 malaria cases
  # Using 'point.in.poly' to return a point spatial object, in this case location of health facilities
  # and estimated population instead of sp::over function, which simply returns 
  # a data frame, with the same no. rows.
  # Argument 'sp = TRUE' returns an sp class object, else returns sf class object
  # Joining the malaria and population dataset using only 'merge' function can't work due to 
  # non-unique columns and differences in row numbers
  
  hospitals_in_catchment <- spatialEco::point.in.poly(malaria_shp, catchment_shp, sp = TRUE) 

  # Add the extracted ID, health facility names and dry season malaria cases to 
  # the health facility catchments (hfc)
  hfc_malaria_shp <- merge(catchment_shp, hospitals_in_catchment, by.x = "DN", by.y = "FID")

  # Convert the shapefile containing malaria data to sf-object
  hfc_malaria_sf <- sf::st_as_sf(hfc_malaria_shp)

  # Tidy the data by dropping columns not needed
  catchment_malaria <- hfc_malaria_sf %>% 
    dplyr::select(-c(coords.x1, coords.x2))

  return(out = catchment_malaria)
}


# Invoking the function ----------------------------------------------------------------------------------
malaria_by_catchment <- assign.malaria.data(malire_new, health_facility_aggr_sf)

```

Assign population data to the health catchment areas

```{r, message=FALSE}

# Helper unction to extract population from WorldPop raster file and assign ---------------------------
# the values to the new catchments.

extract.pop.values <- function(kasungu_pop_raster, catchments){
  # function to extract population from raster file and assign the population to catchments
  # arguments:
  #   kasungu_pop_raster: population raster file clipped to Kasungu district
  #   catchments: shapefile containing the polygons that we wish to use as boundaries
  # returns:
  #   catchments_malaria_pop_sf: sf polygon object containing malaria and population data
  
  # convert from sf to sp
  catchments_sp <- as(catchments, "Spatial")
  
  # Match extent i.e projection
  catchments_sp <- spTransform(catchments_sp, proj4string(kasungu_pop_raster))
  
  # Crop and mask the population raster to exclude Kasungu National Park
  pop_raster_clip <- raster::mask(raster::crop(kasungu_pop_raster, extent(catchments_sp)), catchments_sp)
  
  # Extracting zonal statistics from a population raster layer. 
  # The population raster is a continuous gridded surface layer that has an 
  # estimated population density value to every square in their grid. 
  # The population values are then summed and apportioned to the catchment polygons
  # catchments_malaria_pop <- catchments %>% 
  #   dplyr::mutate(pop = round(raster::extract(pop_raster_clip, catchments, fun = sum, na.rm = TRUE)))
  
  pop_by_catchment <- round(raster::extract(pop_raster_clip, catchments, fun = sum, na.rm = TRUE))
  
  pop_by_catchment_df <-  pop_by_catchment %>%  
  # apply unlist to the lists to have vectors as the list elements
  lapply(unlist) %>% 
  # convert vectors to data.frames
  lapply(as_tibble) %>% 
  # combine the list of data.frames
  bind_rows(., .id = "rowID") %>% 
  # rename the value variable
  dplyr::rename(pop = value)
  
  # Add row ID to column to catchment layer
  catchments$rowID <- 1:nrow(catchments)
  
  # Merge catchment areas and population data 
  pop_by_catchments <- merge(catchments, pop_by_catchment_df, by.x = "rowID", by.y = "rowID")
  
  # Cleaning 'Inf' values
  pop_by_catchments %>% 
    dplyr::mutate_if(is.numeric, list(~na_if(., Inf))) %>% 
    dplyr::mutate_if(is.numeric, list(~na_if(., -Inf)))

  return(out = pop_by_catchments)
  
}

# Invoking the function ---------------------------------------------------------------------------------------
malaria_pop_by_catchment_2017 <- extract.pop.values(kasungu_population_2017, malaria_by_catchment)

malaria_pop_by_catchment_2018 <- extract.pop.values(kasungu_population_2018, malaria_by_catchment)

malaria_pop_by_catchment_2019 <- extract.pop.values(kasungu_population_2019, malaria_by_catchment)

malaria_pop_by_catchment_2020 <- extract.pop.values(kasungu_population_2020, malaria_by_catchment)


```


### View Kasungu population by catchment maps
Estimated total number of people within health facility catchment areas.
```{r, message=FALSE, fig.height = 5, fig.width = 9, fig.cap = 'Fig. 4: Estimated population in Kasungu health facility catchment areas'}
# Helper function to create maps of estimated population by catchment areas --------------------------------
create.population.map <- function(catchment.area, 
                                  variable = "pop", 
                                  title, 
                                  legend.title = "Estimated \n population"){
  # estimated population map
  # catchment.area: estimated population layer from nachulu function
  # variable: variable name (as character, in qoutes)
  # title: map title in quotes
  # legend.title: legend title in qoutes
  # returns:
  #   a tmap-element (plots a map)
  tm_shape(catchment.area)+
    tm_fill(col = variable, 
            breaks = c(0, 13000, 19000, 27000, 35000, 70000, 140000, 200000),
            palette = "YlOrBr",
            title = legend.title)+
    tm_borders(col = "grey",
               lwd = 0.4)+
    tm_layout(legend.position = c(0.75, "bottom"),
              legend.text.size = 0.6,
              legend.title.size = 0.8,
              frame = FALSE)+
    tm_credits(title, 
               position = c(0.3, 0.8), 
               size = 1)
}

# Invoking the function --------------------------------------------------------------------------------
pop_by_catchment_2017 <- create.population.map(malaria_pop_by_catchment_2017, title = "2017")

pop_by_catchment_2018 <- create.population.map(malaria_pop_by_catchment_2018, title = "2018")

pop_by_catchment_2019 <- create.population.map(malaria_pop_by_catchment_2019, title = "2019")

pop_by_catchment_2020 <- create.population.map(malaria_pop_by_catchment_2020, title = "2020")

tmap::tmap_arrange(pop_by_catchment_2017, pop_by_catchment_2018,
                   pop_by_catchment_2019, pop_by_catchment_2020, ncol = 2)


```

### Calculate the expected number of cases for each catchment area 

The `expected` number of dry season malaria cases in catchment *i* are calculated as the observed risk (r) of malaria i.e. the total number of malaria cases in Kasungu district divided by the total population of the district, multiplied by the number of people in the catchment area: $$E_i = \frac{\sum_i O_i}{\sum_i N_i}\times N_i$$

The `expected` number of dry season malaria cases are calculated under the assumption that there is no spatial variation in risk, i.e., no difference in infection rates between the catchment areas.
 
```{r, message=FALSE}

# Calculate expected malaria cases --------------------------------------------------------------
expected_malaria_2017 <- malaria_pop_by_catchment_2017 %>% 
  dplyr::rename(
    observed_2017 = dr_2017,
     pop_2017 = pop) %>% 
  dplyr::mutate(
    expected_2017 = round(sum(observed_2017)/sum(pop_2017, na.rm = TRUE)*pop_2017))

expected_malaria_2018 <- malaria_pop_by_catchment_2018 %>% 
  dplyr::rename(
    observed_2018 = dr_2018,
    pop_2018 = pop) %>% 
  dplyr::mutate(
    expected_2018 = round(sum(observed_2018)/sum(pop_2018, na.rm = TRUE)*pop_2018))

expected_malaria_2019 <- malaria_pop_by_catchment_2019 %>% 
  dplyr::rename(
    observed_2019 = dr_2019,
    pop_2019 = pop) %>%
  dplyr::mutate(
    expected_2019 = round(sum(observed_2019)/sum(pop_2019, na.rm = TRUE)*pop_2019)) 

expected_malaria_2020 <- malaria_pop_by_catchment_2020 %>% 
  dplyr::rename(
    observed_2020 = dr_2020,
    pop_2020 = pop) %>% 
  dplyr::mutate(
    expected_2020 = round(sum(observed_2020)/sum(pop_2020, na.rm = TRUE)*pop_2020))


```

### Calculate the Standardised Morbidity Ratio of malaria incidences for each catchment area 

The `SMR` compares the risk of morbidity in a population of interest with that of a standard population. In this case, our interest is to find out whether the number of dry season malaria cases in each catchment area are greater than we would expect given the malaria rate for the entire Kasungu district.

We do this by comparing what we `observe` (O) with what we would `expect` (E) if the risk of malaria was equal throughout Kasungu. The SMR statistical notation of catchment *i* can be written as follows:
$$SMR_i = \frac{O_i}{E_i}$$

```{r}
# Calculate Standardised Morbidity Ratio (SMR) -------------------------------------
SMR_2017 <- expected_malaria_2017 %>% 
  dplyr::mutate(SMR = round(observed_2017/expected_2017, 1)) %>% 
  dplyr::select(rowID,Names, pop_2017, observed_2017, expected_2017, SMR) 

SMR_2018 <- expected_malaria_2018 %>% 
  dplyr::mutate(SMR = round(observed_2018/expected_2018, 1)) %>% 
  dplyr::select(rowID, Names, pop_2018, observed_2018, expected_2018, SMR) 

SMR_2019 <- expected_malaria_2019 %>% 
  dplyr::mutate(SMR = round(observed_2019/expected_2019, 1)) %>% 
  dplyr::select(rowID, Names, pop_2019, observed_2019, expected_2019, SMR) 

SMR_2020 <- expected_malaria_2020 %>% 
  dplyr::mutate(SMR = round(observed_2020/expected_2020, 1)) %>% 
  dplyr::select(rowID, Names, pop_2020, observed_2020, expected_2020, SMR)


# Create SMR tables ------------------------------------------------------------
SMR_table_2017 <- SMR_2017 %>% 
  dplyr::as_tibble() %>% 
  dplyr::select(-rowID, -geometry) %>% 
  kable %>%
  kableExtra::kable_styling(full_width = FALSE)

SMR_table_2018 <- SMR_2018 %>% 
  dplyr::as_tibble() %>% 
  dplyr::select(-rowID, -geometry) %>% 
  kable %>% 
  kableExtra::kable_styling(full_width = FALSE)

SMR_table_2019 <- SMR_2019 %>% 
  dplyr::as_tibble() %>% 
  dplyr::select(-rowID, -geometry) %>% 
  kable %>% 
  kableExtra::kable_styling(full_width = FALSE)

SMR_table_2020 <- SMR_2020 %>% 
  dplyr::as_tibble() %>% 
  dplyr::select(-rowID, -geometry) %>% 
  kable %>% 
  kableExtra::kable_styling(full_width = FALSE)

SMR_table_2017

SMR_table_2018

SMR_table_2019

SMR_table_2020

```

 
### View observed and expected dry season malaria cases
```{r, message=FALSE, fig.height=10, fig.width=7, fig.cap= 'Fig 5: Observed and expected malaria incidence by health facility catchment area, Kasungu'}
# Helper unction to create maps of observed and expected dry season malaria cases
create.malaria.map <- function(malaria.data, 
                               variable = NA, 
                               title = NA, 
                               legend.title = NA){
  # observed and expected malaria incidence map
  # malaria.data: data frame containing observed and expected malaria cases
  # variable: variable name (as character, in qoutes e.g. "observed")
  # title: map title in quotes
  # legend.title: legend title in qoutes
  # returns:
  #   a tmap-element (plots a map)
  tm_shape(malaria.data)+
    tm_fill(col = variable, 
            breaks = c(0, 500, 1000, 2500, 5000, 10000, 15000, 25000, 35000),
            palette = "YlOrRd",
            title = legend.title)+
    tm_borders(lw = 0.3)+
    tm_layout(legend.position = c(0.75,"bottom"),
              legend.text.size = 0.5,
              legend.title.size = 0.7,
              frame = FALSE)+
    tm_credits(title, 
               position = c(0.2, 0.8), 
               size = 1)
}

# Invoking the function
# 2017 observed and expected malaria cases -------------------------------------
observed_malaria_2017_map <- create.malaria.map(malaria_pop_by_catchment_2017, 
                                                variable = "dr_2017",
                                                title = "2017",
                                                legend.title = "Observed malaria")

expected_malaria_2017_map <- create.malaria.map(expected_malaria_2017,
                                                variable = "expected_2017",
                                                title = "2017",
                                                legend.title = "Expected malaria")

# 2018 observed and expected malaria cases -------------------------------------
observed_malaria_2018_map <- create.malaria.map(malaria_pop_by_catchment_2018,
                                                variable = "dr_2018",
                                                title = "2018",
                                                legend.title = "Observed malaria")

expected_malaria_2018_map <- create.malaria.map(expected_malaria_2018,
                                                variable = "expected_2018",
                                                title = "2018",
                                                legend.title = "Expected malaria")

# 2019 observed and expected malaria cases -------------------------------------
observed_malaria_2019_map <- create.malaria.map(malaria_pop_by_catchment_2019,
                                                variable = "dr_2019",
                                                title = "2019",
                                                legend.title = "Observed malaria")

expected_malaria_2019_map <- create.malaria.map(expected_malaria_2019,
                                                variable = "expected_2019",
                                                title = "2019",
                                                legend.title = "Expected malaria")

# 2020 observed and expected malaria cases -------------------------------------
observed_malaria_2020_map <- create.malaria.map(malaria_pop_by_catchment_2020,
                                                variable = "dr_2020",
                                                title = "2020",
                                                legend.title = "Observed malaria")

expected_malaria_2020_map <- create.malaria.map(expected_malaria_2020,
                                                variable = "expected_2020",
                                                title = "2020",
                                                legend.title = "Expected malaria")

# Layout maps ------------------------------------------------------------------
tmap::tmap_arrange(observed_malaria_2017_map, expected_malaria_2017_map,
                   observed_malaria_2018_map, expected_malaria_2018_map, 
                   observed_malaria_2019_map, expected_malaria_2019_map,
                   observed_malaria_2020_map, expected_malaria_2020_map, ncol = 2)



```

### View SMR by catchment

A ratio greater than 1.0 indicates that more malaria cases have occurred than would have been expected, while a ratio less than 1.0 indicates that less cases have occurred.
```{r, message=FALSE, fig.height = 5, fig.width = 7, fig.cap = 'Fig. 6: Standardised morbidity ratio of malaria by health facility catchment'}

#   max(SMR_2017$SMR)
# [1] 6.93
# > max(SMR_2018$SMR)
# [1] 7.93
# > max(SMR_2019$SMR)
# [1] 7.71
# > max(SMR_2020$SMR)
# [1] 5.68

# Helper function to create maps of SMR by catchment ----------------------------------
create.smr.map <- function(smr.data, 
                           variable = "SMR", 
                           title = NA, 
                           legend.title = "SMR"){
  # SMR by catchment map
  # smr.data: sf polygon object containing SMR by catchment data
  # variable: variable name (as character, in qoutes)
  # title: map title in quotes
  # legend.title: legend title in qoutes
  # returns:
  #   a tmap-element (plots a map)
 
  tm_shape(smr.data)+
    tm_fill(col = variable, 
            breaks = c(0, 0.5, 1, 1.5, 2, 2.5, 5, 8),
            palette = "-magma",
            title = legend.title)+
    tm_borders(lw = 0.3)+
    tm_layout(legend.position = c(0.75,"bottom"),
              legend.text.size = 0.5,
              legend.title.size = 0.7,
              frame = FALSE)+
    tm_credits(title, 
               position = c(0.2, 0.8), 
               size = 1)
}

# Invoking function -------------------------------------------------------------------
SMR_2017_map <- create.smr.map(SMR_2017, title = "2017")

SMR_2018_map <- create.smr.map(SMR_2018, title = "2018")

SMR_2019_map <- create.smr.map(SMR_2019, title = "2019")

SMR_2020_map <- create.smr.map(SMR_2020, title = "2020")

# Layout maps -------------------------------------------------------------------------
tmap::tmap_arrange(SMR_2017_map, SMR_2018_map, SMR_2019_map, SMR_2020_map, ncol = 2)


```

### Calculate the proportion of the catchment population living within 1km, 2km, 3km of water bodies
First, using `st_buffer`, we compute 1km, 2km and 3km buffers around dry season water bodies obtained from LandSat satellite imagery using `TropWet` tool in Google Earth Engine. Then geometry of the buffer features are then combined resulting in resolved internal boundaries to enable extracting population values from WorldPop raster. Finally, we calculate the proportion of people in each catchment area living within water bodies.
```{r, message=FALSE, warning=FALSE}
# Combine and transform TropWet derived waterbody polygons -------------------------------
surface_waterbodies_2017 <- sf::st_as_sf(
  st_cast(
    st_union(
      st_buffer(dryseason_waterbodies_2017, dist = 30)), "POLYGON"))

surface_waterbodies_2018 <- sf::st_as_sf(
  st_cast(
    st_union(
      st_buffer(dryseason_waterbodies_2018, dist = 30)), "POLYGON"))

surface_waterbodies_2019 <- sf::st_as_sf(
  st_cast(
    st_union(
      st_buffer(dryseason_waterbodies_2019, dist = 30)), "POLYGON"))

surface_waterbodies_2020 <- sf::st_as_sf(
  st_cast(
    st_union(
      st_buffer(dryseason_waterbodies_2020, dist = 30)), "POLYGON"))

# Helper function to compute 1km, 2km and 3km buffers around the water bodies ---------------------

create.waterbody.buffer <- function(waterbody, distance, catchment){
  # function for creating buffers around waterbodies
  # arguments:
  #   waterbody:  waterbody shapefile
  #   distance: buffer distance in meters
  #   catchment: catchment area shapefile
  # returns:
  #   buffered waterbodies 
  
  # Create buffers around water bodies
  buffer_radius <- sf::st_buffer(waterbody, distance)
  
  # Dissolve the buffers
  buffer_union <- sf::st_as_sf(st_cast(st_union(buffer_radius),"MULTIPOLYGON"))
  
  # Assign attributes of the 'catchment' to each of the water bodies. 
   buffer_intersect <- sf::st_intersection(buffer_union, catchment)
  
   buffer_intersect_sf <- sf::st_as_sf(buffer_intersect)
   
  # Convert the MULTIPOLYGON object into several POLYGON objects
   buffer_intersect_polygons <- sf::st_cast(
     sf::st_buffer(buffer_intersect_sf,0.0), "MULTIPOLYGON") %>% 
     sf::st_cast("POLYGON")
  
  # Polygons being seen to be in multiple catchments
   sf::st_intersects(buffer_intersect_polygons, catchment)
  
  # Make the assumption that the attribute is constant throughout the geometry
   sf::st_agr(buffer_intersect_polygons) = "constant"
   
   sf::st_agr(catchment) = "constant"
  
  return(out = buffer_intersect_polygons)
}


# Invoking function
# For 2017 TropWet surface water polygons --------------------------------------------------------
buffer_1km_2017 <- create.waterbody.buffer(waterbody = surface_waterbodies_2017, 
                                           distance = 1000, 
                                           catchment = malire_new)

buffer_2km_2017 <- create.waterbody.buffer(waterbody = surface_waterbodies_2017, 
                                           distance = 2000, 
                                           catchment = malire_new)

buffer_3km_2017 <- create.waterbody.buffer(waterbody = surface_waterbodies_2017, 
                                           distance = 3000,
                                           catchment = malire_new)

# For 2018 TropWet surface water polygons --------------------------------------------------------
buffer_1km_2018 <- create.waterbody.buffer(waterbody = surface_waterbodies_2018, 
                                           distance = 1000, 
                                           catchment = malire_new)

buffer_2km_2018 <- create.waterbody.buffer(waterbody = surface_waterbodies_2018, 
                                           distance = 2000, 
                                           catchment = malire_new)

buffer_3km_2018 <- create.waterbody.buffer(waterbody = surface_waterbodies_2018, 
                                           distance = 3000, 
                                           catchment = malire_new)
 
# For 2019 TropWet surface water polygons ------------------------------------------------------
buffer_1km_2019 <- create.waterbody.buffer(waterbody = surface_waterbodies_2019, 
                                           distance = 1000, 
                                           catchment = malire_new)

buffer_2km_2019 <- create.waterbody.buffer(waterbody = surface_waterbodies_2019, 
                                           distance = 2000, 
                                           catchment = malire_new)

buffer_3km_2019 <- create.waterbody.buffer(waterbody = surface_waterbodies_2019, 
                                           distance = 3000, 
                                           catchment = malire_new)

# For 2020 TropWet surface water polygons ------------------------------------------------------
buffer_1km_2020 <- create.waterbody.buffer(waterbody = surface_waterbodies_2020, 
                                           distance = 1000, 
                                           catchment = malire_new)

buffer_2km_2020 <- create.waterbody.buffer(waterbody = surface_waterbodies_2020, 
                                           distance = 2000, 
                                           catchment = malire_new)

buffer_3km_2020 <- create.waterbody.buffer(waterbody = surface_waterbodies_2020, 
                                           distance = 3000, 
                                           catchment = malire_new)

```

View the created waterbody buffers 

```{r, fig.height=9, fig.width=10, fig.cap='Fig 7. Buffers around dry season waterbodies in Kasungu'}

# Map the buffers
create.buffer.map <- function(buffers, boundary = malire_new, title = NA){
  # function for creating buffer map in ggplot
  # arguments:
  #   buffer:  waterbodies buffer polygon layer
  #   boundary: health facility catchment polygons
  #   title: main title
  # returns:
  #   a map-element (plots a map)
  ggplot(data = buffers)+
     geom_sf()+
     geom_sf(data = boundary, 
             fill = NA)+
     theme_void()+
     labs(title = title)
}

# Invoking the function
# For 2017 -------------------------------------------------------------------------------
buffer_1km_2017_map <- create.buffer.map(buffer_1km_2017, title = "2017: 1km Buffers")

buffer_2km_2017_map <- create.buffer.map(buffer_2km_2017, title = "2017: 2km Buffers")

buffer_3km_2017_map <- create.buffer.map(buffer_3km_2017, title = "2017: 3km Buffers")

# For 2018 --------------------------------------------------------------------------------
buffer_1km_2018_map <- create.buffer.map(buffer_1km_2018, title = "2018: 1km Buffers")

buffer_2km_2018_map <- create.buffer.map(buffer_2km_2018, title = "2018: 2km Buffers")

buffer_3km_2018_map <- create.buffer.map(buffer_3km_2018, title = "2018: 3km Buffers")

# For 2019 ---------------------------------------------------------------------------------
buffer_1km_2019_map <- create.buffer.map(buffer_1km_2019, title = "2019: 1km Buffers")

buffer_2km_2019_map <- create.buffer.map(buffer_2km_2019, title = "2019: 2km Buffers")

buffer_3km_2019_map <- create.buffer.map(buffer_3km_2019, title = "2019: 3km Buffers")

# For 2020 --------------------------------------------------------------------------------
buffer_1km_2020_map <- create.buffer.map(buffer_1km_2020, title = "2020: 1km Buffers")

buffer_2km_2020_map <- create.buffer.map(buffer_2km_2020, title = "2020: 2km Buffers")

buffer_3km_2020_map <- create.buffer.map(buffer_3km_2020, title = "2020: 3km Buffers")
 
grid.arrange(buffer_1km_2017_map, buffer_1km_2018_map, buffer_1km_2019_map, buffer_1km_2020_map,
             buffer_2km_2017_map, buffer_2km_2018_map, buffer_2km_2019_map, buffer_2km_2020_map, 
             buffer_3km_2017_map, buffer_3km_2018_map, buffer_3km_2019_map, buffer_3km_2020_map, ncol = 4)


```

Extract the population living within waterbody buffers by catchment area

```{r, message=FALSE, warning=FALSE}

# Helper function to calculate estimated number of people living within waterbody buffers
# in each catchment area
estimate.buffer.pop <- function(catchment.population, buffers, catchment.area){
  
  # Extract population estimates from WorldPop raster
  buffers$buffer_pop <- raster::extract(catchment.population,
                                        buffers, 
                                        fun = sum, 
                                        na.rm = TRUE)
                                               
                                              
  # Find which catchment each polygon belongs to using its centroid - a point dataset 
  # representing the geographic center-points of the polygons 
  buffer_by_catchment <- st_intersection(st_centroid(buffers), catchment.area)
  
  # Notice that the buffer_catchment is comprised of separate POLYGONS (buffer_by_catchment$x). 
  # The first step is to “dissolve” away these POLYGONS into one MULTIPOLYGON. 
  # There is no sf equivalent to the QGIS or ArcMap “dissolve” operation. 
  # Instead we use a combination of group_by and summarize from the dplyr package. 
  # Stats::aggregate from sf package, and dplyr::summarize both do essentially the same.
   buffer_pop_aggregated <- buffer_by_catchment %>% 
     dplyr::group_by(DN) %>%
     dplyr::summarize(
       buffer_pop_aggregated = round(sum(buffer_pop, na.rm = TRUE)))
   
  buffer_pop <- merge(
    catchment.area, st_drop_geometry(
      buffer_pop_aggregated), by = 'DN', all.x = TRUE)
  
  return(out = buffer_pop)
  
}

# Invoking the function and calculating proportion of 
# catchment population living within buffers
# 2017 buffer population -------------------------------------------------------
buffer_pop_1km_2017 <- estimate.buffer.pop(
  kasungu_population_2017, 
  buffer_1km_2017, 
   malaria_pop_by_catchment_2017) %>% 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) %>% 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100)) 

buffer_pop_2km_2017 <- estimate.buffer.pop(
  kasungu_population_2017,
  buffer_2km_2017,
  malaria_pop_by_catchment_2017) %>% 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) %>% 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100)) 

buffer_pop_3km_2017 <- estimate.buffer.pop(
  kasungu_population_2017,
  buffer_3km_2017,
  malaria_pop_by_catchment_2017) %>% 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) %>% 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

# 2018 buffer population -------------------------------------------------------
buffer_pop_1km_2018 <- estimate.buffer.pop(
  kasungu_population_2018,
  buffer_1km_2018,
  malaria_pop_by_catchment_2018) %>% 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) %>% 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

buffer_pop_2km_2018 <- estimate.buffer.pop(
  kasungu_population_2018,
  buffer_2km_2018,
  malaria_pop_by_catchment_2018) %>% 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) %>% 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

buffer_pop_3km_2018 <- estimate.buffer.pop(
  kasungu_population_2018,
  buffer_3km_2018,
  malaria_pop_by_catchment_2018) %>% 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) %>% 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

# 2019 buffer population -------------------------------------------------------
buffer_pop_1km_2019 <- estimate.buffer.pop(
  kasungu_population_2019,
  buffer_1km_2019,
  malaria_pop_by_catchment_2019) %>% 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) %>% 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

buffer_pop_2km_2019 <- estimate.buffer.pop(
  kasungu_population_2019,
  buffer_2km_2019,
  malaria_pop_by_catchment_2019) %>% 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) %>% 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

buffer_pop_3km_2019 <- estimate.buffer.pop(
  kasungu_population_2019,
  buffer_3km_2019,
  malaria_pop_by_catchment_2019) %>% 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) %>% 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

# 2020 buffer population -------------------------------------------------------
buffer_pop_1km_2020 <- estimate.buffer.pop(
  kasungu_population_2020,
  buffer_1km_2020,
  malaria_pop_by_catchment_2020) %>% 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) %>% 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

buffer_pop_2km_2020 <- estimate.buffer.pop(
  kasungu_population_2020,
  buffer_2km_2020,
  malaria_pop_by_catchment_2020) %>% 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) %>% 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

buffer_pop_3km_2020 <- estimate.buffer.pop(
  kasungu_population_2020,
  buffer_3km_2020,
  malaria_pop_by_catchment_2020) %>% 
  dplyr::rename(catchment_pop = pop,
                buffer_pop = buffer_pop_aggregated) %>% 
  dplyr::mutate(
    prop_buffer_catchment_pop = round((buffer_pop/catchment_pop)*100))

```


Mapping proportion of catchment population living within waterbodies

```{r, message=FALSE, warning = FALSE, fig.height = 9, fig.width = 11, fig.cap='Fig 8. Proportion of catchment population living around water bodies'}

# Helper function to create maps of proportion of people living in proximity ----------
# to water bodies in each catchment area
create.pop.proportion.map <- function(pop.data, 
                                      variable = "prop_buffer_catchment_pop", 
                                      title = NA, 
                                      legend.title = NA){
 
  # pop.data: sf polygon object containing proportion of catchment population 
  #           living within water bodies
  # variable: variable name (as character, in qoutes)
  # title: map title in quotes
  # legend.title: legend title in qoutes
  # returns:
  #   a tmap-element (plots a map)
 
  tm_shape(pop.data)+
    tm_fill(col = variable, 
            breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100),
            palette = "YlOrBr",
            title = legend.title)+
    tm_borders(lw = 0.3)+
    tm_layout(legend.position = c(0.8,"bottom"),
              legend.text.size = 0.5,
              legend.title.size = 0.7,
              frame = FALSE)+
    tm_credits(title, 
               position = c(0.25, 0.75), 
               size = 1)
}

# Invoking function 
# 2017 population proportion ---------------------------------------------------
pop_proportion_1km_2017_map <- create.pop.proportion.map(
  buffer_pop_1km_2017, 
  title = "2017",
  legend.title = "Population within \n1km buffers (%)")

pop_proportion_2km_2017_map <- create.pop.proportion.map(
  buffer_pop_2km_2017, 
  title = "2017",
  legend.title = "Population within \n2km buffers (%)")

pop_proportion_3km_2017_map <- create.pop.proportion.map(
  buffer_pop_3km_2017,
  title = "2017",
  legend.title = "Population within \n3km buffers (%)")

# 2018 population proportion ---------------------------------------------------
pop_proportion_1km_2018_map <- create.pop.proportion.map(
  buffer_pop_1km_2018,
  title = "2018",
  legend.title = "Population within \n1km buffers (%)")

pop_proportion_2km_2018_map <- create.pop.proportion.map(
  buffer_pop_2km_2018,
  title = "2018",
  legend.title = "Population within \n2km buffers (%)")

pop_proportion_3km_2018_map <- create.pop.proportion.map(
  buffer_pop_3km_2018,
  title = "2018",
  legend.title = "Population within \n3km buffers (%)")

# 2019 population proportion ---------------------------------------------------
pop_proportion_1km_2019_map <- create.pop.proportion.map(
  buffer_pop_1km_2019,
  title = "2019",
  legend.title = "Population within \n1km buffers (%)")

pop_proportion_2km_2019_map <- create.pop.proportion.map(
  buffer_pop_2km_2019,
  title = "2019",
  legend.title = "Population within \n2km buffers (%)")

pop_proportion_3km_2019_map <- create.pop.proportion.map(
  buffer_pop_3km_2019,
  title = "2019",
  legend.title = "Population within \n3km buffers (%)")

# 2020 population proportion ---------------------------------------------------
pop_proportion_1km_2020_map <- create.pop.proportion.map(
  buffer_pop_1km_2020,
  title = "2020",
  legend.title = "Population within \n1km buffers (%)")

pop_proportion_2km_2020_map <- create.pop.proportion.map(
  buffer_pop_2km_2020,
  title = "2020",
  legend.title = "Population within \n2km buffers (%)")

pop_proportion_3km_2020_map <- create.pop.proportion.map(
  buffer_pop_3km_2020,
  title = "2020",
  legend.title = "Population within \n3km buffers (%)")

# Layout maps ------------------------------------------------------------------
tmap::tmap_arrange(pop_proportion_1km_2017_map, pop_proportion_2km_2017_map, 
                   pop_proportion_3km_2017_map, pop_proportion_1km_2018_map,
                   pop_proportion_2km_2018_map, pop_proportion_3km_2018_map,
                   pop_proportion_1km_2019_map, pop_proportion_2km_2019_map,
                   pop_proportion_3km_2019_map, pop_proportion_1km_2020_map,
                   pop_proportion_2km_2020_map, pop_proportion_3km_2020_map, ncol = 3)
```

### Scatter plots of SMR against the proportion of the catchment population living waterbody buffers

```{r, message=FALSE, warning=FALSE, fig.height=15, fig.width=11, fig.cap='Fig 9. Relationship between standardised morbidity ratio and living near waterbodies'}

# Helper function to tidy and bind the SMR and proportion of -------------------
# buffer-catchment population data frames
tidy.data <- function(smr.df, 
                      proportion.pop.1km, 
                      proprotion.pop.2km,
                      proportion.pop.3km){

# Convert the sf object to dataframes-------------------------------------------
smr_df <- as.data.frame(smr.df) %>% 
  dplyr::select(rowID, Names, SMR)

proportion_pop_1km_df <- as.data.frame(proportion.pop.1km) %>% 
  dplyr::select(rowID, prop_pop_1km = `prop_buffer_catchment_pop`)

proportion_pop_2km_df <- as.data.frame(proprotion.pop.2km)%>% 
  dplyr::select(rowID, prop_pop_2km = `prop_buffer_catchment_pop`)

proportion_pop_3km_df <- as.data.frame(proportion.pop.3km)%>% 
  dplyr::select(rowID, prop_pop_3km = `prop_buffer_catchment_pop`)

# Merge SMR and population data frames -----------------------------------------
combined_1 <- merge(smr_df, proportion_pop_1km_df, by = "rowID", all = TRUE)

combined_2 <- merge(proportion_pop_2km_df, proportion_pop_3km_df)

combined_fully <- merge(combined_1, combined_2, by = "rowID", all = TRUE)

}

# Invoking the function --------------------------------------------------------
smr_pop_2017 <- tidy.data(SMR_2017, buffer_pop_1km_2017, buffer_pop_2km_2017, buffer_pop_3km_2017)

smr_pop_2018 <- tidy.data(SMR_2018, buffer_pop_1km_2018, buffer_pop_2km_2018, buffer_pop_3km_2018)

smr_pop_2019 <- tidy.data(SMR_2019, buffer_pop_1km_2019, buffer_pop_2km_2019, buffer_pop_3km_2019)

smr_pop_2020 <- tidy.data(SMR_2020, buffer_pop_1km_2020, buffer_pop_2km_2020, buffer_pop_3km_2020)

# Helper function to create scatter plots --------------------------------------
create.scatter.plot <- function(smr.pop.df, 
                                independent.var = NA,
                                dependent.var = "SMR",
                                x.label = NA,
                                plot.title = NA){
  
  scatter.plot <- ggpubr::ggscatter(smr.pop.df,          # data frame
                                    x = independent.var, # x-axis variable
                                    y = dependent.var,   # y-axis variable
                                    add = "reg.line",    # Add regression line
                                    conf.int = TRUE,     # Add confidence interval
                                    add.params = list(color = "red",
                                                      fill = "lightgray"),
                                    palette = "jco",     # journal color palette. see ?ggpar
                                    xlab = x.label, 
                                    ylab = "SMR",
                                    title = plot.title)+    
                  ggpubr::stat_cor(label.y = 5)+         # Add correlation coefficient
                  ggpubr::font("title", size = 10, face = "bold")+
                  ggpubr::font("xlab", size = 10)+
                  ggpubr::font("ylab", size = 10)
 
  return(scatter.plot)
  
}

# Invoking function 
# 2017 scatter plots ------------------------------------------------------------

scatter_1km_2017 <- create.scatter.plot(smr_pop_2017, independent.var = "prop_pop_1km",
                                       x.label = "Percentage of catchment population \nliving in 1km buffer",
                                       plot.title = "2017")

scatter_2km_2017 <- create.scatter.plot(smr_pop_2017, independent.var = "prop_pop_2km",
                                        x.label = "Percentage of catchment population \nliving in 2km buffer",
                                        plot.title = "2017")

scatter_3km_2017 <- create.scatter.plot(smr_pop_2017, independent.var = "prop_pop_3km",
                                        x.label = "Percentage of catchment population \nliving in 3km buffer",
                                        plot.title = "2017")

# 2018 scatter plots -----------------------------------------------------------
scatter_1km_2018 <- create.scatter.plot(smr_pop_2018, independent.var = "prop_pop_1km",
                                        x.label = "Percentage of catchment population \nliving in 1km buffer",
                                        plot.title = "2018")

scatter_2km_2018 <- create.scatter.plot(smr_pop_2018, independent.var = "prop_pop_1km",
                                        x.label = "Percentage of catchment population \nliving in 2km buffer",
                                        plot.title = "2018")

scatter



# Create a list of plots -------------------------------------------------------
plot_lists <- list(scatter_1km_2017, scatter_2km_2017, scatter_3km_2017,
                   scatter_1km_2018, scatter_2km_2018, scatter_3km_2019,
                   scatter_1km_2019, scatter_2km_2019, scatter_3km_2019, 
                   scatter_1km_2020, scatter_2km_2020, scatter_3km_2020)

# Arrange the plots ----------------------------------------------------------
grid.arrange(grobs = plot_lists, ncol = 3)

```



 